---
title: "01_duke_weather_tsa"
output: html_document
---

```{r}
library(here)
library(tidyverse)
library(lubridate)
library(nlme)
library(gtsummary)
library(lme4)
library(sf)
library(data.table)
library(stringr)
library(ggplot2)
library(scales)
```

```{r}
#Paths
outage_path <- "data_raw/outage_tracker.csv"                # Duke outage table
lookup_path  <- "data_processed/county_to_uscrn_lookup.csv"  # county->WBANNO mapping
weather_dir  <- "data_processed/NOAA_weather_stations_processed"            # folder with station .rds files
out_merged_rds <- "data_processed/merged_outages_weather_full.rds"
out_merged_csv <- "data_processed/merged_outages_weather_full.csv"

# Matching behavior:
time_tolerance_secs <- 900   # max allowable seconds difference for final nearest-in-time fallback (default 15 min)
```

Helpers
```{r}
# parse fix_duration_estimate like "0d 1h 20m 15s" -> minutes numeric
parse_duration <- function(x) {
  # Remove trailing/leading whitespace
  x <- trimws(x)
  
  # Split on spaces
  parts <- strsplit(x, "\\s+")
  
  # Extract numeric values in fixed order: d h m s
  res <- vapply(parts, function(p) {
    nums <- as.numeric(sub("[^0-9]", "", p))
    if (length(nums) < 4) nums <- c(nums, rep(0, 4 - length(nums)))
    days  <- nums[1]
    hours <- nums[2]
    mins  <- nums[3]
    secs  <- nums[4]
    days * 1440 + hours * 60 + mins + secs / 60
  }, numeric(1))
  
  res
}

# Safe POSIX parse helper (try multiple formats)
safe_parse_datetime <- function(x) {
  dt <- ymd_hms(x, tz = "UTC")
  if (all(is.na(dt))) dt <- ymd_hm(x, tz = "UTC")
  if (all(is.na(dt))) dt <- ymd(x, tz = "UTC")
  dt
}

# Read first non-empty row of each file to extract WBANNO quickly
build_station_index <- function(weather_dir, pattern = "\\.txt$", verbose = TRUE) {
  files <- list.files(weather_dir, full.names = TRUE, pattern = pattern)
  if (length(files) == 0) stop("No weather files found in ", weather_dir)

  idx_list <- lapply(files, function(f) {
    # read the very first non-empty, non-comment line
    line <- readLines(f, n = 1, warn = FALSE)
    if (length(line) == 0 || line == "" || grepl("^#", line)) return(NULL)
    
    # extract just the first number before the first space
    wban <- sub(" .*", "", line)  # everything up to first space
    wban <- trimws(wban)
    
    data.table(WBANNO = wban, file = f)
  })

  idx_dt <- rbindlist(idx_list, use.names = TRUE, fill = TRUE)
  idx_dt[, WBANNO := trimws(WBANNO)]
  idx_dt <- unique(idx_dt, by = "WBANNO")

  if (verbose)
    message("Indexed ", nrow(idx_dt), " unique WBANNOs from ", length(files), " files.")

  return(idx_dt)
}


# Parse a station file fully and return a DT with columns:
# WBANNO, datetime_utc (POSIX), AIR_TEMPERATURE, PRECIPITATION, RELATIVE_HUMIDITY, WIND_1_5, LONGITUDE, LATITUDE
# This assumes a positional column layout consistent with your sample (adjust indices if different).
parse_station_file_to_dt <- function(fpath) {
  dt <- tryCatch(fread(fpath, header = FALSE, sep = "", fill = TRUE, showProgress = FALSE), error = function(e) NULL)
  if (is.null(dt) || nrow(dt) == 0) return(NULL)
  # Ensure column names V1..Vn
  setnames(dt, paste0("V", seq_len(ncol(dt))))
  # Access columns by position with safety checks
  getcol <- function(i) if (i <= ncol(dt)) dt[[i]] else NA
  parsed <- data.table(
    WBANNO = as.character(getcol(1)),
    UTC_DATE = as.character(getcol(2)),          # YYYYMMDD
    UTC_TIME = sprintf("%04d", as.integer(getcol(3))), # HHmm -> force 4 digits
    LONGITUDE = as.numeric(getcol(7)),           # sample mapping; adjust if needed
    LATITUDE = as.numeric(getcol(8)),
    AIR_TEMPERATURE = as.numeric(getcol(9)),
    PRECIPITATION = as.numeric(getcol(10)),
    RELATIVE_HUMIDITY = as.numeric(getcol(16)),
    WIND_1_5 = as.numeric(getcol(22))
  )
  parsed[, datetime_utc := ymd_hm(paste0(UTC_DATE, " ", substr(UTC_TIME,1,2), ":", substr(UTC_TIME,3,4)), tz = "UTC")]
  parsed <- parsed[!is.na(datetime_utc)]
  setkey(parsed, datetime_utc)  # for fast joins/rolls
  return(parsed)
}
```

```{r}
#Read outage table and lookup
outages <- fread(outage_path, stringsAsFactors = FALSE, fill = TRUE)
lookup  <- fread(lookup_path, stringsAsFactors = FALSE)

# Ensure text columns align: trim and unify "County" suffix issues
normalize_name <- function(x) {
  x <- as.character(x) #turn into characters
  x <- str_trim(x) #removing whitespace from start and end of string if any
  x <- str_replace_all(x, "\\s+County$", "")    # drop " County"
  x <- str_to_title(x) #capitalizing the first letter of each word
  x
}

#Updating the county and state in both the outages and lookup files
outages[, county := normalize_name(county)]
outages[, state  := str_trim(state)]
lookup[, county := normalize_name(county)]
lookup[, state  := str_trim(state)]

# Merge to attach WBANNO to each outage (by state & county)
outages <- left_join(outages, lookup, by = c('state', 'county'))

# Parse outage timestamps and duration
outages[, outage_start_dt := safe_parse_datetime(outage_start_estimate)]

# If outage_start_dt is NA but outage_restored + fix_duration_estimate exist, compute start
if (any(is.na(outages$outage_start_dt) & !is.na(outages$outage_restored) & !is.na(outages$fix_duration_estimate))) {
  idx <- which(is.na(outages$outage_start_dt) & !is.na(outages$outage_restored) & !is.na(outages$fix_duration_estimate))
  outages[idx, outage_restored_dt := safe_parse_datetime(outage_restored)]
  outages[idx & !is.na(outage_restored_dt), outage_start_dt := outage_restored_dt - minutes(parse_duration_to_minutes(fix_duration_estimate)[idx & !is.na(outage_restored_dt)])]
}
outages[, fix_duration_minutes := parse_duration(fix_duration_estimate)]
outages[, affected := as.numeric(affected)]
```
I am reformatting data:
Bringing in all weather data
Make a dataframe for each state
Average the weather data from the unique WBANNO by state
Full timeseries for length of raw data: "2024-05-28 18:01:01 UTC" - "2025-09-10 04:30:39 UTC"
Then for each state, bring in the outages
- check 15 min time interval --> is there any outage? --> if so attach the outage data
- check subsequent time intervals to see if the outage is there

```{r}
# ---------- 0) Prepare outages data.table (use your cleaned version) ----------
setDT(outages)
# ensure WBANNO is character
outages[, WBANNO := as.character(WBANNO)]

# Ensure start timestamp parsed already (your pipeline created outage_start_dt)
if (!"outage_start_dt" %in% names(outages)) {
  stop("outages must contain outage_start_dt (POSIX). Run your parsing code first.")
}
if (!inherits(outages$outage_start_dt, "POSIXt")) outages[, outage_start_dt := as.POSIXct(outage_start_dt, tz = "UTC")]

# Determine outage end column
outage_end_col <- if ("outage_end_estimate" %in% names(outages)) "outage_end_estimate" else
                  if ("outage_restored" %in% names(outages)) "outage_restored" else NA
if (is.na(outage_end_col)) {
  # We'll allow NA ends (treated as ongoing) but warn
  message("No outage end column found; outages with NA end will be treated as ongoing until series end.")
} else {
  if (!inherits(outages[[outage_end_col]], "POSIXt")) outages[, (outage_end_col) := as.POSIXct(get(outage_end_col), tz = "UTC")]
}

# compute numeric duration if not present
if (!"fix_duration_minutes" %in% names(outages) && "fix_duration_estimate" %in% names(outages)) {
  outages[, fix_duration_minutes := parse_duration(fix_duration_estimate)]
}

# ---------- 1) Build list of stations from outages (unique WBANNO) ----------
stations_needed <- unique(na.omit(outages$WBANNO))
if (length(stations_needed) == 0) stop("No WBANNOs found in outages. Check your lookup merge.")

# ---------- 2) Helper to read a station RDS file (robust) ----------
read_station_rds_safe <- function(wban, weather_dir, weather_vars) {
  fpath <- file.path(weather_dir, paste0(wban, ".rds"))
  if (!file.exists(fpath)) return(NULL)
  st_raw <- tryCatch(readRDS(fpath), error = function(e) NULL)
  if (is.null(st_raw)) return(NULL)
  st_dt <- as.data.table(st_raw)
  # detect time column (prefer POSIX)
  timecols <- names(st_dt)[sapply(st_dt, inherits, "POSIXt")]
  if (length(timecols) == 0) {
    cand <- names(st_dt)[grepl("date|time|datetime|ts|stamp", names(st_dt), ignore.case = TRUE)]
    if (length(cand) > 0) {
      timecol <- cand[1]
      st_dt[, (timecol) := as.POSIXct(get(timecol), tz = "UTC")]
    } else {
      # fallback: try named 'datetime' or 'date_time'
      fallback <- intersect(c("datetime", "date_time", "obs_time", "time"), names(st_dt))
      if (length(fallback) > 0) {
        timecol <- fallback[1]
        st_dt[, (timecol) := as.POSIXct(get(timecol), tz = "UTC")]
      } else {
        return(NULL)
      }
    }
  } else {
    timecol <- timecols[1]
  }
  setnames(st_dt, timecol, "weather_dt", skip_absent = TRUE)
  st_dt[, weather_dt := as.POSIXct(weather_dt, tz = "UTC")]
  # ensure WBANNO column exists
  if (!"WBANNO" %in% names(st_dt)) st_dt[, WBANNO := wban]
  st_dt[, WBANNO := as.character(WBANNO)]
  # keep only weather_dt, WBANNO and any weather vars that exist
  keepcols <- intersect(c("weather_dt", "WBANNO", weather_vars), names(st_dt))
  st_dt <- st_dt[, ..keepcols]
  # dedupe identical timestamps
  if (nrow(st_dt) > 1 && any(duplicated(st_dt, by = c("WBANNO", "weather_dt")))) {
    setorder(st_dt, WBANNO, weather_dt)
    st_dt <- st_dt[!duplicated(st_dt, by = c("WBANNO", "weather_dt"))]
  }
  return(st_dt)
}

# ---------- 3) Read only the stations we need (memory-savvy) ----------
weather_vars <- c(
  "AIR_TEMPERATURE", "PRECIPITATION",
  "SOLAR_RADIATION", "SR_FLAG", "SURFACE_TEMPERATURE", "ST_TYPE", "ST_FLAG",
  "RELATIVE_HUMIDITY", "RH_FLAG", "SOIL_MOISTURE_5", "SOIL_TEMPERATURE_5",
  "WETNESS", "WET_FLAG", "WIND_1_5", "WIND_FLAG"
)

station_dts <- vector("list", length(stations_needed))
names(station_dts) <- stations_needed
for (w in stations_needed) {
  st_dt <- read_station_rds_safe(w, weather_dir, weather_vars)
  if (!is.null(st_dt)) station_dts[[w]] <- st_dt
}
# combine available station data
stations_all <- rbindlist(Filter(NROW, station_dts), use.names = TRUE, fill = TRUE)
if (nrow(stations_all) == 0) stop("No station data read. Check weather_dir and station files.")

# ---------- 4) Map stations -> state using outages mapping (WBANNO -> state) ----------
# 1) Quick diagnostic: which WBANNOs map to multiple states in outages?
multi_map <- outages[!is.na(WBANNO), .N, by = .(WBANNO, state)]
dup_wban <- multi_map[, .N, by = WBANNO][N > 1]
if (nrow(dup_wban) > 0) {
  message("Found WBANNOs that appear with multiple states (showing up to 20):")
  print(head(multi_map[WBANNO %in% dup_wban$WBANNO][order(WBANNO, -N)], 20))
} else {
  message("No WBANNO -> multiple state mappings detected in outages.")
}

# 2) Build a deterministic 1-row-per-WBANNO mapping: choose the most frequent state for that WBANNO
# (this handles occasional mismatches caused by lookup errors or inconsistent county names)
wban_state_map2 <- outages[!is.na(WBANNO),
                           .N, by = .(WBANNO, state)][order(WBANNO, -N)]
# keep the top (most frequent) state per WBANNO
wban_state_map2 <- wban_state_map2[!duplicated(WBANNO), .(WBANNO, state)]
wban_state_map2[, WBANNO := as.character(WBANNO)]
setkey(wban_state_map2, WBANNO)

# Optional: show how many WBANNOs changed vs naive unique mapping
# (you can inspect multi_map if you want to manually resolve)
message("Mapping contains ", nrow(wban_state_map2), " unique WBANNO -> state rows.")

# 3) Ensure stations_all WBANNO is same type and trimmed of whitespace
stations_all[, WBANNO := as.character(WBANNO)]
stations_all[, WBANNO := trimws(WBANNO)]

# 4) Now merge safely (one-to-many: station rows -> single state)
stations_all <- merge(
  stations_all,
  wban_state_map2,
  by = "WBANNO",
  all.x = TRUE,
  sort = FALSE
)

# ---------- 5) Compute numeric weather variables to average (exclude flags/non-numeric) ----------
numeric_weather_vars <- intersect(names(stations_all), weather_vars)
numeric_weather_vars <- numeric_weather_vars[sapply(stations_all[, ..numeric_weather_vars], is.numeric)]
if (length(numeric_weather_vars) == 0) stop("No numeric weather vars found in station files.")

# state-level average per timewban_state_map <- outages[, .(state = first(state)), by = WBANNO]stamp (mean across stations in same state at that exact weather_dt)
state_weather <- stations_all[!is.na(state),
                              c(list(weather_dt = weather_dt[1]),
                                lapply(.SD, function(x) mean(x, na.rm = TRUE))),
                              by = .(state, weather_dt),
                              .SDcols = numeric_weather_vars]

# ---------- 6) Build full 15-minute timeline for desired range ----------
# candidate vars you want (adjust if names differ)
candidate_weather_vars <- c("AIR_TEMPERATURE","PRECIPITATION","RELATIVE_HUMIDITY",
                            "SOLAR_RADIATION","WIND_1_5","SOIL_MOISTURE_5")

# defensive setup
setDT(state_weather)
state_weather[, timestamp := as.POSIXct(weather_dt, tz = "UTC")]  # if it was weather_dt, rename earlier
setkey(state_weather, state, timestamp)

# Which of the candidate vars are actually present in state_weather?
present_vars <- intersect(candidate_weather_vars, names(state_weather))
missing_vars <- setdiff(candidate_weather_vars, present_vars)
message("Present weather vars: ", paste(present_vars, collapse = ", "))
if (length(missing_vars) > 0) message("Missing/unavailable vars (will be skipped): ", paste(missing_vars, collapse = ", "))

# Use present_vars in the pipeline (if none present, we still make state_ts with timestamp/state)
numeric_weather_vars <- present_vars

# prepare output list
state_ts_list <- vector("list", length(states)); names(state_ts_list) <- states

for (s in states) {
  message("Processing state: ", s)
  # 15-min grid for this state
  dt_ts <- data.table(state = s, timestamp = full_ts)
  dt_ts[, timestamp := as.POSIXct(timestamp, tz = "UTC")]

  # subset state_weather for this state
  st_w <- state_weather[state == s, c("timestamp", numeric_weather_vars), with = FALSE]

  if (nrow(st_w) == 0) {
    message("  no station weather rows for state ", s, " -> leaving all NA for weather")
    state_ts_list[[s]] <- dt_ts
    next
  }

  # rolling join: st_w is X, dt_ts is i. We keep only columns that exist in st_w.
  # roll = time_tolerance_secs will match nearest within +/- tolerance (if NULL use "nearest")
  res <- st_w[dt_ts, on = .(timestamp), roll = time_tolerance_secs]

  # ensure 'state' column present and set order of columns defensively
  res[, state := s]
  desired_cols <- c("state", "timestamp", numeric_weather_vars)
  # keep only those that actually exist in res
  desired_cols <- intersect(desired_cols, names(res))
  setcolorder(res, desired_cols)

  state_ts_list[[s]] <- res
}

# combine
state_ts <- rbindlist(state_ts_list, use.names = TRUE, fill = TRUE)

#TESTTESTTEST
# choose one state (replace "NC" with the state you want)
test_state <- "North Carolina"   # <---- change this if you want a different state

# ----- defensive existence checks -----
if (!exists("state_weather")) stop("state_weather not found in workspace")
if (!exists("stations_all")) message("Warning: stations_all not found (station-level diagnostics will be skipped)")
if (!exists("outages")) message("Warning: outages not found (WBANNO mapping checks will be limited)")
if (!exists("full_ts")) message("Warning: full_ts not found (15-min grid checks will be limited)")

setDT(state_weather)

cat("=== 1) state_weather summary & column names ===\n")
print(names(state_weather))
cat("\nRows for state:", test_state, "->", state_weather[state == test_state, .N], "\n")
print(head(state_weather[state == test_state], 10))

# ----- candidate vars you expect -----
candidate_weather_vars <- c("AIR_TEMPERATURE","PRECIPITATION","RELATIVE_HUMIDITY",
                            "SOLAR_RADIATION","WIND_1_5","SOIL_MOISTURE_5")
present_vars <- intersect(candidate_weather_vars, names(state_weather))
cat("\nCandidate vars present in state_weather:\n"); print(present_vars)
if (length(present_vars) == 0) {
  message("No candidate weather vars present in state_weather. Inspect station files or names.")
} else {
  cat("\nSummary stats for present vars (first few):\n")
  print(sapply(state_weather[state == test_state, ..present_vars], function(x) {
    c(class = class(x)[1], n_non_na = sum(!is.na(x)), min = suppressWarnings(min(x, na.rm=TRUE)), max = suppressWarnings(max(x, na.rm=TRUE)))
  }))
}

# ----- timestamp ranges -----
cat("\n=== 2) timestamp ranges ===\n")
if (exists("full_ts")) {
  cat("Full 15-min grid range:\n"); print(range(full_ts, na.rm = TRUE))
} else cat("full_ts not available\n")
if (state_weather[state == test_state, .N] > 0) {
  cat("state_weather timestamps range for", test_state, ":\n")
  print(range(state_weather[state == test_state, timestamp], na.rm = TRUE))
  cat("First 10 timestamps for that state:\n")
  print(head(state_weather[state == test_state, timestamp], 10))
} else cat("No state_weather rows for this state.\n")

# ----- 15-min grid for test state -----
cat("\n=== 3) build 15-min grid sample ===\n")
if (!exists("full_ts")) {
  # build a small local grid spanning state_weather times if possible
  if (state_weather[state == test_state, .N] > 0) {
    start_grid <- min(state_weather[state == test_state, timestamp], na.rm = TRUE)
    end_grid   <- max(state_weather[state == test_state, timestamp], na.rm = TRUE)
    full_ts_local <- seq(from = start_grid, to = end_grid, by = 15*60)
    cat("Built local grid from state_weather range: ", start_grid, "->", end_grid, "\n")
  } else {
    stop("No full_ts and no state_weather rows to infer a grid.")
  }
} else {
  full_ts_local <- full_ts  # use global
}
dt_ts <- data.table(state = test_state, timestamp = as.POSIXct(full_ts_local, tz = "UTC"))
print(head(dt_ts, 10))

# ----- 4) subset state_weather to st_w for this state and present columns -----
cat("\n=== 4) subset state_weather for rolling join ===\n")
st_w <- state_weather[state == test_state, c("timestamp", present_vars), with = FALSE]
cat("Rows in st_w:", nrow(st_w), "\n")
print(head(st_w, 20))

# ----- 5) try rolling joins (different strategies) -----
time_tolerance_secs <- if (exists("time_tolerance_secs")) time_tolerance_secs else 900
cat("\n=== 5) rolling join: roll = time_tolerance_secs (", time_tolerance_secs, "s ) ===\n")
st_w <- st_w[order(st_w$timestamp)]
res_tol <- left_join(dt_ts, st_w, by = 'timestamp')
  
  st_w[dt_ts, on = .(timestamp)]
if (inherits(res_tol, "try-error")) {
  cat("roll join with tolerance error:\n"); print(res_tol)
} else {
  cat("res_tol rows:", nrow(res_tol), "\n")
  if (nrow(res_tol) > 0 && length(present_vars) > 0) {
    any_non_na <- rowSums(!is.na(res_tol[, ..present_vars])) > 0
    cat("Rows with any non-NA weather (tol):", sum(any_non_na), "of", nrow(res_tol), "\n")
    print("Sample of rows with values:")
    print(head(res_tol[any_non_na], 10))
    print("Sample of rows without values:")
    print(head(res_tol[!any_non_na], 10))
  }
}

cat("\n=== 6) rolling join: roll = 'nearest' (no max-gap) ===\n")
res_nearest <- try(st_w[dt_ts, on = .(timestamp), roll = "nearest"], silent = TRUE)
if (inherits(res_nearest, "try-error")) {
  cat("roll nearest error:\n"); print(res_nearest)
} else {
  cat("res_nearest rows:", nrow(res_nearest), "\n")
  if (nrow(res_nearest) > 0 && length(present_vars) > 0) {
    any_non_na2 <- rowSums(!is.na(res_nearest[, ..present_vars])) > 0
    cat("Rows with any non-NA weather (nearest):", sum(any_non_na2), "of", nrow(res_nearest), "\n")
    print(head(res_nearest[any_non_na2], 10))
  }
}

cat("\n=== 7) LOCF join: roll = Inf (carry-last-forward) ===\n")
res_locf <- try(st_w[dt_ts, on = .(timestamp), roll = Inf], silent = TRUE)
if (inherits(res_locf, "try-error")) {
  cat("roll Inf error:\n"); print(res_locf)
} else {
  if (nrow(res_locf) > 0 && length(present_vars) > 0) {
    any_non_na3 <- rowSums(!is.na(res_locf[, ..present_vars])) > 0
    cat("Rows with any non-NA weather (LOCF):", sum(any_non_na3), "of", nrow(res_locf), "\n")
    print(head(res_locf[any_non_na3], 10))
  }
}

# ----- 6) station-level diagnostics if rolling didn't fill values -----
cat("\n=== 8) station-level diagnostics (stations_all) ===\n")
if (exists("stations_all")) {
  setDT(stations_all)
  cat("stations_all column names:\n"); print(names(stations_all))
  cat("\nNumber of station rows for state:", test_state, "->", stations_all[state == test_state, .N], "\n")
  print(head(stations_all[state == test_state], 20))
  cat("\nCounts per WBANNO for this state:\n")
  print(stations_all[state == test_state, .N, by = WBANNO][order(-N)])
} else {
  message("stations_all not found; skip station-level checks.")
}

# ----- 7) check outages -> WBANNO mapping for this state -----
cat("\n=== 9) outages WBANNO mapping for state ===\n")
if (exists("outages")) {
  setDT(outages)
  print(outages[state == test_state, .N, by = WBANNO][order(-N)])
  example_wban <- outages[state == test_state & !is.na(WBANNO), as.character(WBANNO)][1]
  cat("Example WBANNO for this state from outages:", example_wban, "\n")
  if (!is.na(example_wban)) {
    fpath <- file.path(weather_dir, paste0(example_wban, ".rds"))
    cat("Corresponding station file path:", fpath, "\n")
    if (file.exists(fpath)) {
      st_raw <- readRDS(fpath)
      cat("station RDS structure:\n"); print(str(st_raw))
      if (is.data.frame(st_raw)) print(head(as.data.table(st_raw), 20))
    } else cat("Station RDS file not found for that WBANNO.\n")
  }
} else {
  message("outages not present; cannot inspect WBANNO mapping.")
}

# ----- final note -----
cat("\n=== DONE diagnostics for state:", test_state, "===\n")
cat("If none of the rolling joins produced non-NA values, check:\n",
    " - present_vars (columns are not present or all NA),\n",
    " - stations_all has no rows for this state, or\n",
    " - station RDS files use different column names than expected.\n")

# ---------- 8) Attach outages to each 15-min slot per state (non-equi join) ----------
# ensure types
state_ts[, timestamp := as.POSIXct(timestamp, tz = "UTC")]
outages_dt[, start := as.POSIXct(start, tz = "UTC")]
outages_dt[, end   := as.POSIXct(end,   tz = "UTC")]
outages_dt[, outage_identifier := as.character(outage_identifer)]

# vectorized non-equi join: X = outages_dt, i = state_ts
matched_all <- outages_dt[state_ts, 
                          on = .(state, start <= timestamp, end >= timestamp),
                          nomatch = 0,
                          allow.cartesian = TRUE,
                          .(state = i.state, timestamp = i.timestamp, outage_identifer = outage_identifer)]

agg_all <- matched_all[, .(
  n_active_outages = uniqueN(outage_identifer),
  outage_ids = paste(sort(unique(outage_identifer)), collapse = ";")
), by = .(state, timestamp)]

state_ts_with_outages <- merge(state_ts, agg_all, by = c("state", "timestamp"), all.x = TRUE, sort = FALSE)
state_ts_with_outages[is.na(n_active_outages), `:=`(n_active_outages = 0L, outage_ids = NA_character_)]
state_ts_with_outages[, outage_present := n_active_outages > 0L]

# ---------- 9) Compute contiguous outage runs per state (persistence) ----------
state_ts_with_outages[, outage_seq_id := rleid(outage_present), by = state]
runs <- state_ts_with_outages[, .(start_ts = min(timestamp), end_ts = max(timestamp), n_slots = .N, present = first(outage_present)), by = .(state, outage_seq_id)]
outage_runs <- runs[present == TRUE]

# ---------- 10) Save or return results ----------
# Save if you want:
saveRDS(state_ts_with_outages, file.path("data_processed/", "full_weather_outages.rds"))

# Quick inspect
message("Completed. Example rows:")
print(state_ts_with_outages[1:10])
message("Example outage runs:")
print(outage_runs[1:10])

# Return objects to workspace
state_ts_with_outages -> final_state_ts_with_outages
outage_runs -> final_outage_runs
```

# Exploratory Plots

```{r}
dt <- copy(state_ts_with_outages)
setDT(dt)

# candidate weather vars you might want to plot
candidate_weather_vars <- c("AIR_TEMPERATURE","PRECIPITATION","RELATIVE_HUMIDITY",
                            "SOLAR_RADIATION","WIND_1_5","SOIL_MOISTURE_5")

# pick only the candidate vars that actually exist and are numeric
cand <- intersect(candidate_weather_vars, names(dt))
numeric_weather_vars <- cand[vapply(dt[, ..cand], is.numeric, logical(1))]
message("Will attempt to plot numeric weather vars: ", paste(numeric_weather_vars, collapse = ", "))

if (length(numeric_weather_vars) == 0) stop("No numeric weather variables found in data.")

# create plot dir
plot_dir <- "plots/state_plots"
if (!dir.exists(plot_dir)) dir.create(plot_dir, recursive = TRUE)

states <- sort(unique(dt$state))

for (s in states) {
  message("Processing state: ", s)
  sd <- dt[state == s, c("timestamp", "n_active_outages", numeric_weather_vars), with = FALSE]

  if (nrow(sd) == 0) {
    message("  no rows for state ", s, " -> skipping")
    next
  }

  # ensure timestamp is POSIXct
  sd[, timestamp := as.POSIXct(timestamp, tz = "UTC")]

  # normalize every numeric_weather_var to a new column <var>_norm (guaranteed creation)
  norm_vars <- character(0)
  for (v in numeric_weather_vars) {
    col_norm <- paste0(v, "_norm")
    colvals <- sd[[v]]
    if (all(is.na(colvals))) {
      sd[[col_norm]] <- NA_real_
    } else {
      mn <- min(colvals, na.rm = TRUE)
      mx <- max(colvals, na.rm = TRUE)
      if (is.infinite(mn) || is.infinite(mx) || mx == mn) {
        # constant or degenerate: set to 0.5 so it appears in middle
        sd[[col_norm]] <- rep(0.5, nrow(sd))
      } else {
        sd[[col_norm]] <- (colvals - mn) / (mx - mn)
      }
    }
    norm_vars <- c(norm_vars, col_norm)
  }

  # scale outages to 0-1 (per state)
  max_out <- sd[, max(n_active_outages, na.rm = TRUE)]
  if (!is.finite(max_out) || max_out == 0) {
    sd[, n_out_scaled := 0]
  } else {
    sd[, n_out_scaled := n_active_outages / max_out]
  }

  # keep only norm_vars that actually exist in sd (defensive)
  norm_vars <- norm_vars[norm_vars %in% names(sd)]
  if (length(norm_vars) == 0) {
    message("  no normalized weather columns for state ", s, " -> skipping")
    next
  }

  # melt for plotting
  m <- melt(sd,
            id.vars = c("timestamp", "n_active_outages", "n_out_scaled"),
            measure.vars = norm_vars,
            variable.name = "weather_var",
            value.name = "value_norm",
            variable.factor = FALSE)

  # tidy weather_var names
  m[, weather_var := sub("_norm$", "", weather_var)]

  # build plot
  p <- ggplot() +
    geom_col(data = unique(m[, .(timestamp, n_out_scaled)]),
             aes(x = timestamp, y = n_out_scaled),
             fill = "grey50", alpha = 0.35, width = 900) +   # 900 sec = 15 min
    geom_line(data = m, aes(x = timestamp, y = value_norm, color = weather_var), size = 0.6, na.rm = TRUE) +
    scale_color_brewer(type = "qual", palette = "Set1", name = "Weather") +
    labs(title = paste0("State: ", s),
         subtitle = "Normalized weather (lines) and outages (bars = # outages per 15-min)",
         x = "Time", y = "Normalized weather (0-1)") +
    theme_minimal() +
    theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

  # add secondary axis mapping 0-1 back to outage count
  if (is.finite(max_out) && max_out > 0) {
    p <- p + scale_y_continuous(sec.axis = sec_axis(~ . * max_out, name = "Outages per 15-min slot", labels = comma))
  } else {
    p <- p + scale_y_continuous(sec.axis = sec_axis(~ . * max_out, name = "Outages per 15-min slot"))
  }

  # save plot
  fname <- file.path(plot_dir, paste0("state_plot_", gsub("[^A-Za-z0-9_-]", "_", s), ".png"))
  ggsave(fname, plot = p, width = 14, height = 5, dpi = 150)
  message("  saved: ", fname)
}

message("All done. Plots in: ", normalizePath(plot_dir))
```


