---
title: "01_saidi_saifi_caidi_duke"
output: html_document
---

```{r package install}
library(tidyverse)
library(lubridate)
library(here)
library(sf)
library(dplyr)
library(ggplot2)
library(readxl)
library(plotly)
library(shiny)
library(data.table)
library(tidycensus)
```

```{r data pull}
#outage_data_raw <- fread(here("data_raw/outage_tracker.csv"))
#saveRDS(outage_data_raw, here("data_processed/outage_tracker.rds"))
outage_data_raw <- readRDS(here("data_processed/outage_tracker.rds"))  #saving it in a binary format so it's easier to work with

#Filling in county and state data for entries missing this but have convex hull data
outage_data_sf <- outage_data_raw %>% 
  filter(is.na(county) | is.na(state)) %>% #rows missing county/state
```

SAIDI (System Average Interruption Duration Index)
= ∑(customers affected × outage duration) / total customers served

SAIFI (System Average Interruption Frequency Index)
= ∑(customers affected) / total customers served

CAIDI (Customer Average Interruption Duration Index)
= SAIFI / SAIDI

Ideal to use convex hulls, but unfortunately unsure of customers served in a convex hull, so going to stick with county data.

```{r tidycensus population data}
# population data by county for all the states

#Apply the API key
census_api_key("b5acdf7f791fa15dda8522426afc7a2e6e922c02", install = TRUE, overwrite = TRUE)
readRenviron("~/.Renviron")
```

```{r}
# From ACS (5-year estimates, includes households, income, etc.)
county_pop <- get_acs(
  geography = "county",
  variables = "B01003_001E",  # Total population
  year = 2022,
  state = c('FL', 'GA', 'IL', 'IN', 'KY', 'NC', 'OH', 'SC', 'VA')
)

county_pop_clean <- county_pop %>% 
  separate(NAME, into = c("county", "state"), sep = ", ") %>%
  rename(population = estimate) %>% 
  select("GEOID", "county", "state", "population")

outage_pop_combined <- left_join(outage_data_raw, county_pop_clean, by = c('county', 'state'))
```

Some of the rows are missing state and county data, so information is gleaned from convex hull data:

