---
title: "02_duke_weather_analysis"
output: html_document
---

```{r}
library(here)
library(tidyverse)
library(lubridate)
library(nlme)
library(gtsummary)
library(lme4)
library(sf)
library(data.table)
library(stringr)
```

```{r}
#Paths
outage_path <- "data_raw/outage_tracker.csv"                # Duke outage table
lookup_path  <- "data_processed/county_to_uscrn_lookup.csv"  # county->WBANNO mapping
weather_dir  <- "data_processed/NOAA_weather_stations_processed"            # folder with station .rds files
out_merged_rds <- "data_processed/merged_outages_weather_full.rds"
out_merged_csv <- "data_processed/merged_outages_weather_full.csv"

# Matching behavior:
time_tolerance_secs <- 900   # max allowable seconds difference for final nearest-in-time fallback (default 15 min)
```

#Process of combining duke outages and weather data

Helper Functions:
```{r}
# parse fix_duration_estimate like "0d 1h 20m 15s" -> minutes numeric
parse_duration <- function(x) {
  # Remove trailing/leading whitespace
  x <- trimws(x)
  
  # Split on spaces
  parts <- strsplit(x, "\\s+")
  
  # Extract numeric values in fixed order: d h m s
  res <- vapply(parts, function(p) {
    nums <- as.numeric(sub("[^0-9]", "", p))
    if (length(nums) < 4) nums <- c(nums, rep(0, 4 - length(nums)))
    days  <- nums[1]
    hours <- nums[2]
    mins  <- nums[3]
    secs  <- nums[4]
    days * 1440 + hours * 60 + mins + secs / 60
  }, numeric(1))
  
  res
}

# Safe POSIX parse helper (try multiple formats)
safe_parse_datetime <- function(x) {
  dt <- ymd_hms(x, tz = "UTC")
  if (all(is.na(dt))) dt <- ymd_hm(x, tz = "UTC")
  if (all(is.na(dt))) dt <- ymd(x, tz = "UTC")
  dt
}

# Read first non-empty row of each file to extract WBANNO quickly
build_station_index <- function(weather_dir, pattern = "\\.txt$", verbose = TRUE) {
  files <- list.files(weather_dir, full.names = TRUE, pattern = pattern)
  if (length(files) == 0) stop("No weather files found in ", weather_dir)

  idx_list <- lapply(files, function(f) {
    # read the very first non-empty, non-comment line
    line <- readLines(f, n = 1, warn = FALSE)
    if (length(line) == 0 || line == "" || grepl("^#", line)) return(NULL)
    
    # extract just the first number before the first space
    wban <- sub(" .*", "", line)  # everything up to first space
    wban <- trimws(wban)
    
    data.table(WBANNO = wban, file = f)
  })

  idx_dt <- rbindlist(idx_list, use.names = TRUE, fill = TRUE)
  idx_dt[, WBANNO := trimws(WBANNO)]
  idx_dt <- unique(idx_dt, by = "WBANNO")

  if (verbose)
    message("Indexed ", nrow(idx_dt), " unique WBANNOs from ", length(files), " files.")

  return(idx_dt)
}


# Parse a station file fully and return a DT with columns:
# WBANNO, datetime_utc (POSIX), AIR_TEMPERATURE, PRECIPITATION, RELATIVE_HUMIDITY, WIND_1_5, LONGITUDE, LATITUDE
# This assumes a positional column layout consistent with your sample (adjust indices if different).
parse_station_file_to_dt <- function(fpath) {
  dt <- tryCatch(fread(fpath, header = FALSE, sep = "", fill = TRUE, showProgress = FALSE), error = function(e) NULL)
  if (is.null(dt) || nrow(dt) == 0) return(NULL)
  # Ensure column names V1..Vn
  setnames(dt, paste0("V", seq_len(ncol(dt))))
  # Access columns by position with safety checks
  getcol <- function(i) if (i <= ncol(dt)) dt[[i]] else NA
  parsed <- data.table(
    WBANNO = as.character(getcol(1)),
    UTC_DATE = as.character(getcol(2)),          # YYYYMMDD
    UTC_TIME = sprintf("%04d", as.integer(getcol(3))), # HHmm -> force 4 digits
    LONGITUDE = as.numeric(getcol(7)),           # sample mapping; adjust if needed
    LATITUDE = as.numeric(getcol(8)),
    AIR_TEMPERATURE = as.numeric(getcol(9)),
    PRECIPITATION = as.numeric(getcol(10)),
    RELATIVE_HUMIDITY = as.numeric(getcol(16)),
    WIND_1_5 = as.numeric(getcol(22))
  )
  parsed[, datetime_utc := ymd_hm(paste0(UTC_DATE, " ", substr(UTC_TIME,1,2), ":", substr(UTC_TIME,3,4)), tz = "UTC")]
  parsed <- parsed[!is.na(datetime_utc)]
  setkey(parsed, datetime_utc)  # for fast joins/rolls
  return(parsed)
}
```


```{r}
#Read outage table and lookup
outages <- fread(outage_path, stringsAsFactors = FALSE, fill = TRUE)
lookup  <- fread(lookup_path, stringsAsFactors = FALSE)

# Ensure text columns align: trim and unify "County" suffix issues
normalize_name <- function(x) {
  x <- as.character(x) #turn into characters
  x <- str_trim(x) #removing whitespace from start and end of string if any
  x <- str_replace_all(x, "\\s+County$", "")    # drop " County"
  x <- str_to_title(x) #capitalizing the first letter of each word
  x
}

#Updating the county and state in both the outages and lookup files
outages[, county := normalize_name(county)]
outages[, state  := str_trim(state)]
lookup[, county := normalize_name(county)]
lookup[, state  := str_trim(state)]

# Merge to attach WBANNO to each outage (by state & county)
outages <- left_join(outages, lookup, by = c('state', 'county'))

# Parse outage timestamps and duration
outages[, outage_start_dt := safe_parse_datetime(outage_start_estimate)]

# If outage_start_dt is NA but outage_restored + fix_duration_estimate exist, compute start
if (any(is.na(outages$outage_start_dt) & !is.na(outages$outage_restored) & !is.na(outages$fix_duration_estimate))) {
  idx <- which(is.na(outages$outage_start_dt) & !is.na(outages$outage_restored) & !is.na(outages$fix_duration_estimate))
  outages[idx, outage_restored_dt := safe_parse_datetime(outage_restored)]
  outages[idx & !is.na(outage_restored_dt), outage_start_dt := outage_restored_dt - minutes(parse_duration_to_minutes(fix_duration_estimate)[idx & !is.na(outage_restored_dt)])]
}
outages[, fix_duration_minutes := parse_duration(fix_duration_estimate)]
outages[, affected := as.numeric(affected)]
```
for each row, we want to match WBANNO with the respective file name in NOAA_weather_stations_processed.rds.
Then we want to match outage_start_dt between both files and then add on all the weather variables("AIR_TEMPERATURE", "PRECIPITATION",
  "SOLAR_RADIATION", "SR_FLAG", "SURFACE_TEMPERATURE", "ST_TYPE", "ST_FLAG",
  "RELATIVE_HUMIDITY", "RH_FLAG", "SOIL_MOISTURE_5", "SOIL_TEMPERATURE_5",
  "WETNESS", "WET_FLAG", "WIND_1_5", "WIND_FLAG") to the outages file. Except I want to replicate each outage three times, and have it with the weather 1 hour before, 30 minutes before, and the current weather. And have a tag that says which time frame it is.

```{r}
# list of weather variables to attach (must match columns inside station RDS files)
weather_vars <- c(
  "AIR_TEMPERATURE", "PRECIPITATION",
  "SOLAR_RADIATION", "SR_FLAG", "SURFACE_TEMPERATURE", "ST_TYPE", "ST_FLAG",
  "RELATIVE_HUMIDITY", "RH_FLAG", "SOIL_MOISTURE_5", "SOIL_TEMPERATURE_5",
  "WETNESS", "WET_FLAG", "WIND_1_5", "WIND_FLAG"
)

# join options
roll_option <- "nearest"   # "nearest" or remove for exact
maxgap_secs <- NULL        # e.g. 30*60  (30 minutes) or NULL for no limit

# turn outages into data table
outages_dt <- as.data.table(outages)

# add id and create expanded (3 rows per outage)
outages_dt[, outage_id := .I]
expanded <- outages_dt[, .(outage_id, WBANNO, outage_start_dt)][
  , .(
      tag = c("1_hour_before", "30_min_before", "current"),
      target_dt = outage_start_dt + minutes(c(-60, -30, 0))
    ),
  by = .(outage_id, WBANNO, outage_start_dt)
]

expanded_ordered <- expanded[order(expanded$outage_start_dt)]

# Collects all unique station ids
stations <- unique(expanded_ordered$WBANNO)
#Makes a list of length of number of stations you have. And each station in the list will contain all the weather values we are looking for
out_list <- vector("list", length(stations))
#makes it so that we can call each list by its station WBANNO and not just the index in the list
names(out_list) <- stations

# Pre-alloc
out_list <- vector("list", length(stations))

# Optional: whether to print diagnostic messages
VERBOSE <- TRUE

for (i in seq_along(stations)) {
  st <- stations[i]
  st_file <- file.path(weather_dir, paste0(st, ".rds"))
  exp_sub <- copy(expanded_ordered[WBANNO == st])   # per-station expanded subset
  if (nrow(exp_sub) == 0) {
    if (VERBOSE) message("No expanded rows for station: ", st)
    out_list[[i]] <- NULL
    next
  }

  # If station file missing -> produce rows with NA weather columns
  if (!file.exists(st_file)) {
    if (VERBOSE) message("Station file missing for ", st, " -> filling NAs")
    tmp <- exp_sub[, .(outage_id, WBANNO, outage_start_dt, tag, target_dt)]
    for (v in weather_vars) tmp[, (v) := NA]
    tmp[, weather_dt := as.POSIXct(NA)]
    out_list[[i]] <- tmp
    next
  }

  # read station file and coerce to data.table
  st_raw <- readRDS(st_file)
  st_dt  <- as.data.table(st_raw)

  # --- find a time column in the station table if not explicitly provided ---
  # user previously set timecol <- 'outage_start_dt' but station files may differ.
  timecol <- "outage_start_dt"
  if (!timecol %in% names(st_dt)) {
    # try to detect a datetime-like column
    cand <- names(st_dt)[grepl("date|time|datetime|ts|stamp", names(st_dt), ignore.case = TRUE)]
    # prefer POSIXt columns if present
    posix_cols <- names(st_dt)[sapply(st_dt, inherits, what = "POSIXt")]
    if (length(posix_cols) > 0) {
      timecol <- posix_cols[1]
    } else if (length(cand) > 0) {
      timecol <- cand[1]
    } else {
      timecol <- NULL
    }
  }

  if (is.null(timecol)) {
    warning(sprintf("No datetime-like column found in station file %s; filling NA for this station.", st_file))
    tmp <- exp_sub[, .(outage_id, WBANNO, outage_start_dt, tag, target_dt)]
    for (v in weather_vars) tmp[, (v) := NA]
    tmp[, weather_dt := as.POSIXct(NA)]
    out_list[[i]] <- tmp
    next
  }

  # convert timecol to POSIXct if needed and rename to weather_dt
  if (!inherits(st_dt[[timecol]], "POSIXt")) {
    st_dt[, (timecol) := as.POSIXct(get(timecol), tz = "UTC")]
  }
  # rename to 'weather_dt' (skip if already that name)
  setnames(st_dt, old = timecol, new = "weather_dt", skip_absent = TRUE)

  # keep only relevant columns: weather_dt + requested weather_vars
  keepcols <- intersect(c("weather_dt", weather_vars), names(st_dt))
  if (!"weather_dt" %in% keepcols) {
    warning("Station file ", st_file, " does not contain a parsed weather datetime. Filling NA for this station.")
    tmp <- exp_sub[, .(outage_id, WBANNO, outage_start_dt, tag, target_dt)]
    for (v in weather_vars) tmp[, (v) := NA]
    tmp[, weather_dt := as.POSIXct(NA)]
    out_list[[i]] <- tmp
    next
  }
  st_dt <- st_dt[, ..keepcols]

  # tag station id (safe and ensures consistent type)
  st_dt[, WBANNO := as.character(st)]

  # deduplicate identical timestamps per station if present
  if (any(duplicated(st_dt, by = c("WBANNO", "weather_dt")))) {
    if (VERBOSE) message("Deduplicating station rows for ", st, " on (WBANNO, weather_dt)")
    # Simple strategy: keep the first row for each (WBANNO, weather_dt).
    # If you prefer a summary (mean/median), replace this block with an aggregation.
    st_dt <- st_dt[order(WBANNO, weather_dt)]
    st_dt <- st_dt[!duplicated(st_dt, by = c("WBANNO", "weather_dt"))]
  }

  # cast types and set keys for robust joining
  st_dt[, WBANNO := as.character(WBANNO)]
  setkey(st_dt, WBANNO, weather_dt)

  # ensure exp_sub keys and types are character for WBANNO and POSIXct for target_dt
  exp_sub[, WBANNO := as.character(WBANNO)]
  if (!inherits(exp_sub[["target_dt"]], "POSIXt")) {
    exp_sub[, target_dt := as.POSIXct(target_dt, tz = "UTC")]
  }
  setkey(exp_sub, WBANNO, target_dt)

  # ---- IMPORTANT: use st_dt[exp_sub, ...] so result rows follow exp_sub (one result per exp_sub row) ----
  if (is.null(maxgap_secs)) {
    joined <- st_dt[exp_sub,
                    on = .(WBANNO, weather_dt = target_dt),
                    mult = "first",
                    roll = roll_option,
                    nomatch = NA]
  } else {
    joined <- st_dt[exp_sub,
                    on = .(WBANNO, weather_dt = target_dt),
                    mult = "first",
                    roll = roll_option,
                    maxgap = maxgap_secs,
                    nomatch = NA]
  }

  # joined now has columns from exp_sub plus columns from st_dt (including weather_dt and weather_vars)
  # make sure outage_id and other meta from exp_sub are present
  if (!"outage_id" %in% names(joined)) {
    stop("After join, 'outage_id' not found. Check exp_sub content for station ", st)
  }

  # choose columns to keep: preserve outage metadata from exp_sub and available weather vars
  avail_wvars <- intersect(weather_vars, names(joined))
  cols_keep <- c("outage_id", "WBANNO", "outage_start_dt", "tag", "target_dt", "weather_dt", avail_wvars)
  cols_keep <- intersect(cols_keep, names(joined))  # keep only those that exist
  joined <- joined[, ..cols_keep]

  # Defensive: ensure the row order matches exp_sub order (optional)
  # If you want the same row ordering as exp_sub, you can reorder by outage_id or target_dt
  # joined <- joined[order(outage_id, target_dt)]

  out_list[[i]] <- joined
}

# combine results (drop any NULL entries)
weather_matched <- rbindlist(Filter(NROW, out_list), use.names = TRUE, fill = TRUE)


# merge back original outage meta (other columns)
outage_meta_cols <- setdiff(names(outages_dt), c("outage_start_dt","outage_id"))
final <- merge(
  weather_matched,
  outages_dt[, c("outage_id", outage_meta_cols), with = FALSE],
  by = "outage_id",
  all.x = TRUE
)
```
Renaming and reordering columns:
```{r}
# Start by renaming one clean WBANNO column and dropping duplicates
final[, WBANNO := WBANNO.x]                # keep WBANNO.x as the main station id
final[, c("WBANNO.x", "WBANNO.y") := NULL] # drop both redundant columns

# identify your weather columns
weather_vars <- c(
  "AIR_TEMPERATURE", "PRECIPITATION", "SOLAR_RADIATION", "SR_FLAG",
  "SURFACE_TEMPERATURE", "ST_TYPE", "ST_FLAG", "RELATIVE_HUMIDITY",
  "RH_FLAG", "SOIL_MOISTURE_5", "SOIL_TEMPERATURE_5", "WETNESS",
  "WET_FLAG", "WIND_1_5", "WIND_FLAG"
)

# build desired column order
before_weather <- c(
  "outage_id", "WBANNO", "outage_start_dt", "tag", "target_dt",
  "outage_identifer", "device_lat", "device_lon", "block_fips",
  "convex_hull", "jurisdiction", "origin", "state", "county",
  "affected", "cause", "outage_start_estimate", "outage_end_estimate",
  "fix_duration_estimate", "outage_restored",
  "nearest_station_distance_m", "station_within_50km", "fix_duration_minutes"
)

# keep only columns that actually exist, then append weather vars in order
cols_to_keep <- c(intersect(before_weather, names(final)), intersect(weather_vars, names(final)))

# reorder columns
setcolorder(final, cols_to_keep)

setnames(final, old = "tag", new = "weather_lag")

# inspect
colnames(final)

final_ordered <- final[order(final$outage_start_dt)]
saveRDS(final, file.path("data_processed/", "duke_weather_join.rds"))
```
