---
title: "02_duke_weather_analysis"
output: html_document
---

I am collecting reliability metrics from 9 states, sampling different weather quantities in each state
num_outages ~  UTC_DATE + UTC_TIME + AIR_TEMPERATURE + PRECIPITATION + RELATIVE_HUMIDITY + LONGITUDE + LATITUDE, random = ~1 | State/County, data = all of data

duration_outage ~  UTC_DATE + UTC_TIME + AIR_TEMPERATURE + PRECIPITATION + RELATIVE_HUMIDITY + LONGITUDE + LATITUDE, random = ~1 | State/County, data = all of data

```{r}
library(here)
library(tidyverse)
library(lubridate)
library(nlme)
library(gtsummary)
library(lme4)
library(sf)
library(data.table)
library(stringr)
```

```{r}
#Paths
outage_path <- "data_raw/outage_tracker.csv"                # Duke outage table
lookup_path  <- "data_processed/county_to_uscrn_lookup.csv"  # county->WBANNO mapping
weather_dir  <- "data_raw/NOAA_weather_stations"            # folder with station .txt files
out_merged_rds <- "data_processed/merged_outages_weather_full.rds"
out_merged_csv <- "data_processed/merged_outages_weather_full.csv"

# Matching behavior:
time_tolerance_secs <- 900   # max allowable seconds difference for final nearest-in-time fallback (default 15 min)
# If you want only exact matches and no fallback, set time_tolerance_secs <- 0
# -------------------------------------------------------
```

Helper Functions:
```{r}
# parse fix_duration_estimate like "0d 1h 20m 15s" -> minutes numeric
parse_duration <- function(x) {
  # Remove trailing/leading whitespace
  x <- trimws(x)
  
  # Split on spaces
  parts <- strsplit(x, "\\s+")
  
  # Extract numeric values in fixed order: d h m s
  res <- vapply(parts, function(p) {
    nums <- as.numeric(sub("[^0-9]", "", p))
    if (length(nums) < 4) nums <- c(nums, rep(0, 4 - length(nums)))
    days  <- nums[1]
    hours <- nums[2]
    mins  <- nums[3]
    secs  <- nums[4]
    days * 1440 + hours * 60 + mins + secs / 60
  }, numeric(1))
  
  res
}

# Safe POSIX parse helper (try multiple formats)
safe_parse_datetime <- function(x) {
  dt <- ymd_hms(x, tz = "UTC")
  if (all(is.na(dt))) dt <- ymd_hm(x, tz = "UTC")
  if (all(is.na(dt))) dt <- ymd(x, tz = "UTC")
  dt
}

# Read first non-empty row of each file to extract WBANNO quickly
build_station_index <- function(weather_dir, pattern = "\\.txt$", verbose = TRUE) {
  files <- list.files(weather_dir, full.names = TRUE, pattern = pattern)
  if (length(files) == 0) stop("No weather files found in ", weather_dir)

  idx_list <- lapply(files, function(f) {
    # read the very first non-empty, non-comment line
    line <- readLines(f, n = 1, warn = FALSE)
    if (length(line) == 0 || line == "" || grepl("^#", line)) return(NULL)
    
    # extract just the first number before the first space
    wban <- sub(" .*", "", line)  # everything up to first space
    wban <- trimws(wban)
    
    data.table(WBANNO = wban, file = f)
  })

  idx_dt <- rbindlist(idx_list, use.names = TRUE, fill = TRUE)
  idx_dt[, WBANNO := trimws(WBANNO)]
  idx_dt <- unique(idx_dt, by = "WBANNO")

  if (verbose)
    message("Indexed ", nrow(idx_dt), " unique WBANNOs from ", length(files), " files.")

  return(idx_dt)
}


# Parse a station file fully and return a DT with columns:
# WBANNO, datetime_utc (POSIX), AIR_TEMPERATURE, PRECIPITATION, RELATIVE_HUMIDITY, WIND_1_5, LONGITUDE, LATITUDE
# This assumes a positional column layout consistent with your sample (adjust indices if different).
parse_station_file_to_dt <- function(fpath) {
  dt <- tryCatch(fread(fpath, header = FALSE, sep = "", fill = TRUE, showProgress = FALSE), error = function(e) NULL)
  if (is.null(dt) || nrow(dt) == 0) return(NULL)
  # Ensure column names V1..Vn
  setnames(dt, paste0("V", seq_len(ncol(dt))))
  # Access columns by position with safety checks
  getcol <- function(i) if (i <= ncol(dt)) dt[[i]] else NA
  parsed <- data.table(
    WBANNO = as.character(getcol(1)),
    UTC_DATE = as.character(getcol(2)),          # YYYYMMDD
    UTC_TIME = sprintf("%04d", as.integer(getcol(3))), # HHmm -> force 4 digits
    LONGITUDE = as.numeric(getcol(7)),           # sample mapping; adjust if needed
    LATITUDE = as.numeric(getcol(8)),
    AIR_TEMPERATURE = as.numeric(getcol(9)),
    PRECIPITATION = as.numeric(getcol(10)),
    RELATIVE_HUMIDITY = as.numeric(getcol(16)),
    WIND_1_5 = as.numeric(getcol(22))
  )
  parsed[, datetime_utc := ymd_hm(paste0(UTC_DATE, " ", substr(UTC_TIME,1,2), ":", substr(UTC_TIME,3,4)), tz = "UTC")]
  parsed <- parsed[!is.na(datetime_utc)]
  setkey(parsed, datetime_utc)  # for fast joins/rolls
  return(parsed)
}
```


```{r}
#Read outage table and lookup
outages <- fread(outage_path, stringsAsFactors = FALSE, fill = TRUE)
lookup  <- fread(lookup_path, stringsAsFactors = FALSE)

# Ensure text columns align: trim and unify "County" suffix issues
normalize_name <- function(x) {
  x <- as.character(x) #turn into characters
  x <- str_trim(x) #removing whitespace from start and end of string if any
  x <- str_replace_all(x, "\\s+County$", "")    # drop " County"
  x <- str_to_title(x) #capitalizing the first letter of each word
  x
}

#Updating the county and state in both the outages and lookup files
outages[, county := normalize_name(county)]
outages[, state  := str_trim(state)]
lookup[, county := normalize_name(county)]
lookup[, state  := str_trim(state)]

# Merge to attach WBANNO to each outage (by state & county)
outages <- left_join(outages, lookup, by = c('state', 'county'))
#Add 0 in front of 3758 and 3728 so weather station lookup works later
outages <- outages %>%
  mutate(WBANNO = ifelse(WBANNO %in% c("3758", "3728"),
                         paste0("0", WBANNO),
                         WBANNO))

# Parse outage timestamps and duration
outages[, outage_start_dt := safe_parse_datetime(outage_start_estimate)]

# If outage_start_dt is NA but outage_restored + fix_duration_estimate exist, compute start
if (any(is.na(outages$outage_start_dt) & !is.na(outages$outage_restored) & !is.na(outages$fix_duration_estimate))) {
  idx <- which(is.na(outages$outage_start_dt) & !is.na(outages$outage_restored) & !is.na(outages$fix_duration_estimate))
  outages[idx, outage_restored_dt := safe_parse_datetime(outage_restored)]
  outages[idx & !is.na(outage_restored_dt), outage_start_dt := outage_restored_dt - minutes(parse_duration_to_minutes(fix_duration_estimate)[idx & !is.na(outage_restored_dt)])]
}
outages[, fix_duration_minutes := parse_duration(fix_duration_estimate)]
outages[, affected := as.numeric(affected)]

#Build station index and determine which WBANNOs we need to read
station_index <- build_station_index(weather_dir)
needed_wban <- unique(na.omit(outages$WBANNO))
#This checks to see if any weather stations are missing
missing_index <- setdiff(needed_wban, station_index$WBANNO)
if (length(missing_index) > 0) {
  warning("These WBANNOs from lookup are missing station files: ", paste(head(missing_index, 10), collapse = ", "),
          if (length(missing_index) > 10) paste0(", ... (", length(missing_index), " total)"))
}
station_index_needed <- station_index[WBANNO %in% needed_wban]
```

Need to process weather station files
```{r}
parse_station_file_to_dt <- function(path, tz = "UTC", verbose = FALSE) {
  if (verbose) message("Parsing: ", basename(path))

  # Try to read with fread sep="" (splits on runs of whitespace)
  dt_raw <- tryCatch(
    fread(path, header = FALSE, sep = "", fill = TRUE, showProgress = FALSE),
    error = function(e) {
      warning("fread failed on ", path, " : ", e$message)
      return(NULL)
    }
  )
  if (is.null(dt_raw) || nrow(dt_raw) == 0) {
    if (verbose) message("No rows read for ", basename(path))
    return(NULL)
  }

  # Expected NOAA columns
  expected_cols <- c(
    "WBANNO","UTC_DATE","UTC_TIME","LST_DATE","LST_TIME","CRX_VN",
    "LONGITUDE","LATITUDE","AIR_TEMPERATURE","PRECIPITATION",
    "SOLAR_RADIATION","SR_FLAG","SURFACE_TEMPERATURE","ST_TYPE",
    "ST_FLAG","RELATIVE_HUMIDITY","RH_FLAG","SOIL_MOISTURE_5",
    "SOIL_TEMPERATURE_5","WETNESS","WET_FLAG","WIND_1_5","WIND_FLAG"
  )
  n_expected <- length(expected_cols)

  # If fread produced one column (lines not split), or fewer columns than expected, split manually
  if (ncol(dt_raw) == 1L) {
    if (verbose) message("fread produced 1 column; splitting lines on whitespace for ", basename(path))
    lines <- as.character(dt_raw[[1]])
    # drop blank and comment lines at start
    lines <- lines[lines != ""]
    if (length(lines) == 0) return(NULL)
    # tokenise
    toks <- strsplit(lines, "\\s+")
    maxcols <- max(vapply(toks, length, integer(1)))
    # build matrix (pad with NA)
    mat <- t(vapply(toks, function(x) c(x, rep(NA_character_, maxcols - length(x))), character(maxcols)))
    df <- as.data.frame(mat, stringsAsFactors = FALSE)
    dt <- as.data.table(df)
  } else {
    # dt_raw already splitted into multiple columns, use it
    dt <- copy(dt_raw)
  }

  # If dt has more columns than expected, trim; if fewer, pad with NA
  ncols <- ncol(dt)
  if (ncols < n_expected) {
    # pad with NA columns
    for (i in (ncols + 1):n_expected) dt[[i]] <- NA_character_
  } else if (ncols > n_expected) {
    # If extra trailing columns exist, keep only the first expected columns
    # But sometimes files include extra trailing blank column; trim quietly
    dt <- dt[, 1:n_expected, with = FALSE]
  }
  setnames(dt, expected_cols)

  # Clean textual columns
  dt[, WBANNO := trimws(as.character(WBANNO))]
  dt[, UTC_DATE := trimws(as.character(UTC_DATE))]
  dt[, UTC_TIME := trimws(as.character(UTC_TIME))]
  dt[, LST_DATE := trimws(as.character(LST_DATE))]
  dt[, LST_TIME := trimws(as.character(LST_TIME))]

  # Ensure HHmm strings (pad left)
  dt[, UTC_TIME := ifelse(is.na(UTC_TIME) | UTC_TIME == "", NA_character_, sprintf("%04s", UTC_TIME))]
  dt[, LST_TIME := ifelse(is.na(LST_TIME) | LST_TIME == "", NA_character_, sprintf("%04s", LST_TIME))]

  # Numeric conversion for common weather columns
  numeric_cols <- c(
    "LONGITUDE","LATITUDE","AIR_TEMPERATURE","PRECIPITATION",
    "SOLAR_RADIATION","SURFACE_TEMPERATURE","RELATIVE_HUMIDITY",
    "SOIL_MOISTURE_5","SOIL_TEMPERATURE_5","WETNESS","WIND_1_5"
  )
  for (nc in intersect(numeric_cols, names(dt))) {
    # suppressWarnings: some rows have -9999 or non-numeric; will become NA
    dt[[nc]] <- suppressWarnings(as.numeric(dt[[nc]]))
  }

  # Build POSIX datetimes
  dt[, start_datetime_utc := as.POSIXct(NA)]
  ok_start <- !is.na(UTC_DATE) & UTC_DATE != "" & !is.na(UTC_TIME) & UTC_TIME != ""
  if (any(ok_start)) {
    dt[ok_start, start_datetime_utc := ymd_hm(
      paste0(UTC_DATE[ok_start], " ",
             substr(UTC_TIME[ok_start], 1, 2), ":",
             substr(UTC_TIME[ok_start], 3, 4)),
      tz = tz)]
  }

  dt[, end_datetime_utc := as.POSIXct(NA)]
  ok_end <- !is.na(LST_DATE) & LST_DATE != "" & !is.na(LST_TIME) & LST_TIME != ""
  if (any(ok_end)) {
    dt[ok_end, end_datetime_utc := ymd_hm(
      paste0(LST_DATE[ok_end], " ",
             substr(LST_TIME[ok_end], 1, 2), ":",
             substr(LST_TIME[ok_end], 3, 4)),
      tz = tz)]
  }

  # Optionally drop text date/time columns to save memory:
  # dt[, c("UTC_DATE","UTC_TIME","LST_DATE","LST_TIME") := NULL]

  # Ensure WBANNO is character (leading zeros preserved)
  dt[, WBANNO := as.character(WBANNO)]

  # set key for joins if possible
  if (all(c("WBANNO","start_datetime_utc") %in% names(dt))) setkeyv(dt, c("WBANNO","start_datetime_utc"))

  if (verbose) {
    message(sprintf("Parsed %s: rows=%d, start_na=%d, end_na=%d",
                    basename(path), nrow(dt), sum(is.na(dt$start_datetime_utc)), sum(is.na(dt$end_datetime_utc))))
  }

  return(dt[])
}
```

```{r}
# For each needed station, read file, and store in a list keyed by WBANNO (read once)
station_data_list <- list()
for (i in seq_len(nrow(station_index_needed))) {
  w <- station_index_needed$WBANNO[i]
  f <- station_index_needed$file[i]
  message("Reading station ", w, " from ", basename(f), " ...")
  
  sd <- parse_station_file_to_dt(f)
  if (!is.null(sd)) station_data_list[[w]] <- sd
}
```
# GAMs

Worth trying out GAMs (Generalized Additive Models) next because I want longitude and latitude to capture smooth spatial trends rather than discrete group effects, random effects won’t fully capture that.

# Later

Add socioecnomic indicators

population + median income + percent of household with no cars + percent households spending more than 30% of income on housing costs + winter + summer , random = ~1 | State, data = all_of_ data