---
title: "02_duke_weather_analysis"
output: html_document
---

I am collecting reliability metrics from 9 states, sampling different weather quantities in each state
num_outages ~  UTC_DATE + UTC_TIME + AIR_TEMPERATURE + PRECIPITATION + RELATIVE_HUMIDITY + LONGITUDE + LATITUDE, random = ~1 | State/County, data = all of data

duration_outage ~  UTC_DATE + UTC_TIME + AIR_TEMPERATURE + PRECIPITATION + RELATIVE_HUMIDITY + LONGITUDE + LATITUDE, random = ~1 | State/County, data = all of data

```{r}
library(here)
library(tidyverse)
library(lubridate)
library(nlme)
library(gtsummary)
library(lme4)
library(sf)
library(data.table)
library(stringr)
```

```{r}
#Paths
outage_path <- "data_raw/outage_tracker.csv"                # Duke outage table
lookup_path  <- "data_processed/county_to_uscrn_lookup.csv"  # county->WBANNO mapping
weather_dir  <- "data_processed/NOAA_weather_stations_processed"            # folder with station .txt files
out_merged_rds <- "data_processed/merged_outages_weather_full.rds"
out_merged_csv <- "data_processed/merged_outages_weather_full.csv"

# Matching behavior:
time_tolerance_secs <- 900   # max allowable seconds difference for final nearest-in-time fallback (default 15 min)
# If you want only exact matches and no fallback, set time_tolerance_secs <- 0
# -------------------------------------------------------
```

#Process of combining duke outages and weather data

Helper Functions:
```{r}
# parse fix_duration_estimate like "0d 1h 20m 15s" -> minutes numeric
parse_duration <- function(x) {
  # Remove trailing/leading whitespace
  x <- trimws(x)
  
  # Split on spaces
  parts <- strsplit(x, "\\s+")
  
  # Extract numeric values in fixed order: d h m s
  res <- vapply(parts, function(p) {
    nums <- as.numeric(sub("[^0-9]", "", p))
    if (length(nums) < 4) nums <- c(nums, rep(0, 4 - length(nums)))
    days  <- nums[1]
    hours <- nums[2]
    mins  <- nums[3]
    secs  <- nums[4]
    days * 1440 + hours * 60 + mins + secs / 60
  }, numeric(1))
  
  res
}

# Safe POSIX parse helper (try multiple formats)
safe_parse_datetime <- function(x) {
  dt <- ymd_hms(x, tz = "UTC")
  if (all(is.na(dt))) dt <- ymd_hm(x, tz = "UTC")
  if (all(is.na(dt))) dt <- ymd(x, tz = "UTC")
  dt
}

# Read first non-empty row of each file to extract WBANNO quickly
build_station_index <- function(weather_dir, pattern = "\\.txt$", verbose = TRUE) {
  files <- list.files(weather_dir, full.names = TRUE, pattern = pattern)
  if (length(files) == 0) stop("No weather files found in ", weather_dir)

  idx_list <- lapply(files, function(f) {
    # read the very first non-empty, non-comment line
    line <- readLines(f, n = 1, warn = FALSE)
    if (length(line) == 0 || line == "" || grepl("^#", line)) return(NULL)
    
    # extract just the first number before the first space
    wban <- sub(" .*", "", line)  # everything up to first space
    wban <- trimws(wban)
    
    data.table(WBANNO = wban, file = f)
  })

  idx_dt <- rbindlist(idx_list, use.names = TRUE, fill = TRUE)
  idx_dt[, WBANNO := trimws(WBANNO)]
  idx_dt <- unique(idx_dt, by = "WBANNO")

  if (verbose)
    message("Indexed ", nrow(idx_dt), " unique WBANNOs from ", length(files), " files.")

  return(idx_dt)
}


# Parse a station file fully and return a DT with columns:
# WBANNO, datetime_utc (POSIX), AIR_TEMPERATURE, PRECIPITATION, RELATIVE_HUMIDITY, WIND_1_5, LONGITUDE, LATITUDE
# This assumes a positional column layout consistent with your sample (adjust indices if different).
parse_station_file_to_dt <- function(fpath) {
  dt <- tryCatch(fread(fpath, header = FALSE, sep = "", fill = TRUE, showProgress = FALSE), error = function(e) NULL)
  if (is.null(dt) || nrow(dt) == 0) return(NULL)
  # Ensure column names V1..Vn
  setnames(dt, paste0("V", seq_len(ncol(dt))))
  # Access columns by position with safety checks
  getcol <- function(i) if (i <= ncol(dt)) dt[[i]] else NA
  parsed <- data.table(
    WBANNO = as.character(getcol(1)),
    UTC_DATE = as.character(getcol(2)),          # YYYYMMDD
    UTC_TIME = sprintf("%04d", as.integer(getcol(3))), # HHmm -> force 4 digits
    LONGITUDE = as.numeric(getcol(7)),           # sample mapping; adjust if needed
    LATITUDE = as.numeric(getcol(8)),
    AIR_TEMPERATURE = as.numeric(getcol(9)),
    PRECIPITATION = as.numeric(getcol(10)),
    RELATIVE_HUMIDITY = as.numeric(getcol(16)),
    WIND_1_5 = as.numeric(getcol(22))
  )
  parsed[, datetime_utc := ymd_hm(paste0(UTC_DATE, " ", substr(UTC_TIME,1,2), ":", substr(UTC_TIME,3,4)), tz = "UTC")]
  parsed <- parsed[!is.na(datetime_utc)]
  setkey(parsed, datetime_utc)  # for fast joins/rolls
  return(parsed)
}
```


```{r}
#Read outage table and lookup
outages <- fread(outage_path, stringsAsFactors = FALSE, fill = TRUE)
lookup  <- fread(lookup_path, stringsAsFactors = FALSE)

# Ensure text columns align: trim and unify "County" suffix issues
normalize_name <- function(x) {
  x <- as.character(x) #turn into characters
  x <- str_trim(x) #removing whitespace from start and end of string if any
  x <- str_replace_all(x, "\\s+County$", "")    # drop " County"
  x <- str_to_title(x) #capitalizing the first letter of each word
  x
}

#Updating the county and state in both the outages and lookup files
outages[, county := normalize_name(county)]
outages[, state  := str_trim(state)]
lookup[, county := normalize_name(county)]
lookup[, state  := str_trim(state)]

# Merge to attach WBANNO to each outage (by state & county)
outages <- left_join(outages, lookup, by = c('state', 'county'))
#Add 0 in front of 3758 and 3728 so weather station lookup works later
outages <- outages %>%
  mutate(WBANNO = ifelse(WBANNO %in% c("3758", "3728"),
                         paste0("0", WBANNO),
                         WBANNO))

# Parse outage timestamps and duration
outages[, outage_start_dt := safe_parse_datetime(outage_start_estimate)]

# If outage_start_dt is NA but outage_restored + fix_duration_estimate exist, compute start
if (any(is.na(outages$outage_start_dt) & !is.na(outages$outage_restored) & !is.na(outages$fix_duration_estimate))) {
  idx <- which(is.na(outages$outage_start_dt) & !is.na(outages$outage_restored) & !is.na(outages$fix_duration_estimate))
  outages[idx, outage_restored_dt := safe_parse_datetime(outage_restored)]
  outages[idx & !is.na(outage_restored_dt), outage_start_dt := outage_restored_dt - minutes(parse_duration_to_minutes(fix_duration_estimate)[idx & !is.na(outage_restored_dt)])]
}
outages[, fix_duration_minutes := parse_duration(fix_duration_estimate)]
outages[, affected := as.numeric(affected)]
```
for each row, we want to match WBANNO with the respective file name in NOAA_weather_stations_processed.rds.
Then we want to match outage_start_dt between both files and then add on all the weather variables("AIR_TEMPERATURE", "PRECIPITATION",
  "SOLAR_RADIATION", "SR_FLAG", "SURFACE_TEMPERATURE", "ST_TYPE", "ST_FLAG",
  "RELATIVE_HUMIDITY", "RH_FLAG", "SOIL_MOISTURE_5", "SOIL_TEMPERATURE_5",
  "WETNESS", "WET_FLAG", "WIND_1_5", "WIND_FLAG") to the outages file

```{r}

```

# GAMs

Worth trying out GAMs (Generalized Additive Models) next because I want longitude and latitude to capture smooth spatial trends rather than discrete group effects, random effects won’t fully capture that.

# Later

Add socioecnomic indicators

population + median income + percent of household with no cars + percent households spending more than 30% of income on housing costs + winter + summer , random = ~1 | State, data = all_of_ data