---
title: "03_duke_weather_analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(here)
library(tidyverse)
library(lubridate)
library(nlme)
library(gtsummary)
library(lme4)
library(sf)
library(data.table)
library(stringr)
```

Read in table that has all of the duke outages and weather data
```{r}
duke_weather_join <- readRDS("./data_processed/duke_weather_join.rds")
```

Adding columns for the following:
- Month
- Season
- Hour
- Weekend/weekday

```{r}
#Add a month column here
duke_weather_updated <- duke_weather_join %>% 
  mutate('hour' = hour(outage_start_dt)) %>% 
  mutate('day_number' = wday(outage_start_dt)) %>% 
  mutate('month_number' = month(outage_start_dt)) %>% 
  mutate('year_number' = year(outage_start_dt)) %>% 
  mutate('season' = ifelse(month_number %in% c(12,1,2),'Winter',
                           ifelse(month_number %in% c(3,4,5), 'Spring', 
                                  ifelse(month_number %in% c(6,7,8), 'Summer', 'Fall'
                                         )
                                  )
                           )
         ) %>% 
  mutate('day_of_week' = ifelse(day_number %in% c(1,7), 1, 0)) #0 for weekday, 1 for weekend
  #Note that wday's default is 1 for Sunday and 7 for Saturday

#order by date
duke_weather_updated <- duke_weather_updated[order(duke_weather_updated$outage_start_dt)]
```

# Exploratory
Checking distributions of numeric variables to see if anything needs to be transformed

```{r}
numeric_vars <- duke_weather_updated %>%
  select(where(is.numeric)) %>%                    
  select(-any_of(c("device_lat","device_lon","outage_id","WBANNO", "outage_identifer", "affected", "block_fips")))

numeric_vars_dis <- numeric_vars %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  theme_bw() +
  labs(title = "Distribution of Numeric Variables")

numeric_vars_dis

```

Variables to log transform since they are very right-skewed:
- fix_duration_minutes
- SOLAR_RADIATION
- WIND_1_5

```{r}
#Log transformation of the columns
duke_weather_transformed <- duke_weather_updated %>%
  mutate(
    fix_duration_minutes_log = log1p(fix_duration_minutes),
    SOLAR_RADIATION_log      = log1p(SOLAR_RADIATION),
    WIND_1_5_log             = log1p(WIND_1_5)
  )
 
numeric_vars_transformed <- duke_weather_transformed %>%
  select(where(is.numeric)) %>%                    
  select(-any_of(c("device_lat","device_lon","outage_id","WBANNO", "outage_identifer", "affected", "block_fips")))

numeric_vars_transformed_dis <- numeric_vars_transformed %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ variable, scales = "free") +
  theme_bw() +
  labs(title = "Distribution of Numeric Variables")

numeric_vars_transformed_dis
```

# GLM
I am collecting reliability metrics from 9 states, sampling different weather quantities in each state
fix_duration_estimate ~  outage_start_dt + AIR_TEMPERATURE + PRECIPITATION + RELATIVE_HUMIDITY + SOLAR_RADIATION_log + WIND_1_5_log + SOIL_MOISTURE_5 + device_lat + device_lon, random = ~1 | State/County, data = all of data I want to try first seeing if it also depends on season

then want to try if it depends on day_of_week (or can i keep both in the same equation)

duration_outage ~  UTC_DATE + UTC_TIME + AIR_TEMPERATURE + PRECIPITATION + RELATIVE_HUMIDITY + LONGITUDE + LATITUDE, random = ~1 | State/County, data = all of data

```{r}
#fitting a linear mixed-effects model
# variable list

dt <- as.data.table(duke_weather_transformed)

# Step 1 — define numvars in your global environment
numvars <- c("AIR_TEMPERATURE","PRECIPITATION",
             "RELATIVE_HUMIDITY","SOLAR_RADIATION_log","WIND_1_5_log",
             "SOIL_MOISTURE_5","device_lat","device_lon")

# Step 2 — confirm numvars is visible
print(numvars)

# Step 3 — now this will work
sds <- sapply(dt[, ..numvars], function(x) sd(x, na.rm = TRUE))
print(sds)

cor_mat <- cor(dt[, ..numvars], use = "pairwise.complete.obs")
print(round(cor_mat, 3))

duration_lme1 <- gls(
  fix_duration_minutes_log ~ AIR_TEMPERATURE + PRECIPITATION +
    RELATIVE_HUMIDITY + SOLAR_RADIATION_log + WIND_1_5_log +
    device_lat + device_lon,
  data = duke_weather_transformed,
  na.action = na.omit
)

summary(duration_lme1)
```


# GAMs

Worth trying out GAMs (Generalized Additive Models) next because I want longitude and latitude to capture smooth spatial trends rather than discrete group effects, random effects won’t fully capture that.

# Later

Add socioecnomic indicators

population + median income + percent of household with no cars + percent households spending more than 30% of income on housing costs + winter + summer , random = ~1 | State, data = all_of_ data