---
title: "TADS_vegetation"
output: html_document
---

```{r}
library(tidyverse)
library(lubridate)
library(here)
library(sf)
library(leaflet)
library(mapview); mapviewOptions(fgb = FALSE)
library(dplyr)
library(ggplot2)
library(maps)
library(readxl)
library(plotly)
library(shiny)
library(stringr)
#install.packages('leaflet.extras')
library(leaflet.extras)
library(data.table)

#Disable on-the-fly projections
sf::sf_use_s2(FALSE)

#New package to install
#install.packages("climate")
library(climate)
#install.packages("noaa")
library(noaa)
```
```{r}
#need this extra for handling time zones
tz_mapping <- c(
  "EST" = "America/New_York",      # Eastern Standard Time
  "EPT" = "America/New_York",      # Eastern Prevailing Time
  "CST" = "America/Chicago",       # Central Standard Time
  "CPT" = "America/Chicago",       # Central Prevailing Time
  "MPT" = "America/Denver",        # Mountain Prevailing Time
  "MST" = "America/Denver",        # Mountain Standard Time
  "PPT" = "America/Los_Angeles",   # Pacific Prevailing Time
  "PST" = "America/Los_Angeles",   # Pacific Standard Time
  "APT" = "America/Halifax",       # Atlantic Prevailing Time
  "AST" = "America/Halifax",       # Atlantic Standard Time
  "GMT" = "UTC"                    # GMT (Greenwich Mean Time)
)

#PUT REGION HERE: TRE, MRO, NPCC, RFC, SERC, WECC
NERC_region <- 'SERC'
```


I would like to see how many transmission outage events are caused by vegetation divided by total amount for each day in the dataset.

```{r}
#Pulling AC Line inventory
setwd("/Users/JChuang/Library/CloudStorage/GoogleDrive-jchuang@lbl.gov/Shared drives/project_support/outages/data/April_download")
TADS_2013_2021_raw <- read_excel("TADS_AC Circuits_2013_2021_20230711.xlsx", sheet = "AC_Inv_Detail")
TADS_2022_2023_raw <- read_excel("TADS_AC Circuits_2022_20230519.xlsx", sheet = "AC_Inv_Detail")
TADS_2023_2024_raw <- read_excel("TADS_AC Circuits_2023_20240514.xlsx", sheet = "AC_Inv_Detail")
setwd("/Users/JChuang/Library/CloudStorage/GoogleDrive-jchuang@lbl.gov/Shared drives/project_support/outages/data/May_download")
TADS_2024_2025_raw <- read_excel("TADS_AC Circuits_2024_20250506.xlsx", sheet = "AC_Inv_Detail")
TADS_AC <- rbind(TADS_2013_2021_raw, TADS_2022_2023_raw, TADS_2023_2024_raw, TADS_2024_2025_raw)

# Ensure date columns are in Date format
TADS_AC$InServiceDate <- as.Date(TADS_AC$InServiceDate)
TADS_AC$RetirementDate <- as.numeric(TADS_AC$RetirementDate)
TADS_AC$RetirementDate[is.na(TADS_AC$RetirementDate)] <- NA
TADS_AC$RetirementDate <- as.Date("1899-12-30") + TADS_AC$RetirementDate 

inventory_filtered <- TADS_AC# %>% 
 # filter(RegionCode == NERC_region)

#Unique voltage classes
voltage_classes <- unique(inventory_filtered$VoltageClassCodeName)

#Calculating AC inventory counts for a single year
get_AC_counter_df_for_year <- function(year){
  #Filter for the specific year in interest
  inv_year <- inventory_filtered %>% 
    filter(ReportingYearNbr == year)
  
  #Creating data grid for the year
  date_df <- tibble(date = seq(as.Date(paste0(year, "-01-01")), as.Date(paste0(year,"-12-31")), by = "day"))
  
  #create date-voltage grid
  date_voltage_grid <- expand_grid(date=date_df$date, VoltageClassCodeName = voltage_classes)
  
  #cross join and filter for in-service
  date_voltage_cross <- date_voltage_grid %>% 
    rename(JoinDate = date, JoinVoltage = VoltageClassCodeName) %>% 
    mutate(key = 1) %>% 
    inner_join(inv_year %>% mutate(key = 1), by = "key") %>% 
    filter(
      InServiceDate <= JoinDate,
      is.na(RetirementDate) | RetirementDate > JoinDate,
      VoltageClassCodeName == JoinVoltage
    )
  
  # Daily count
    AC_daily_counts <- date_voltage_cross %>%
      group_by(JoinDate, JoinVoltage) %>%
      summarise(total_AC_Inv = n(), .groups = "drop") %>%
      rename(date = JoinDate, VoltageClassCodeName = JoinVoltage)
  
    # Expand to hourly
    AC_counter_df <- AC_daily_counts %>%
      crossing(hour = 0:23) %>%
      mutate(datetime_utc = as.POSIXct(paste(date, sprintf("%02d:00:00", hour)), tz = "UTC")) %>%
      select(datetime_utc, VoltageClassCodeName, total_AC_Inv) %>%
      mutate(ReportingYearNbr = year)  # keep track of year
  
    return(AC_counter_df)
}

years <- unique(inventory_filtered$ReportingYearNbr)
AC_inventory_counter <- map_dfr(years, get_AC_counter_df_for_year)

ggplot(AC_inventory_counter, aes(x=datetime_utc, y = total_AC_Inv, color = VoltageClassCodeName)) +
  geom_line() +
  labs(
    title = "Hourly count of inventory",
    x = "Month", 
    y = "Inventory Count",
    color = "Voltage Class"  
  ) +
  theme_minimal() 

AC_inventory_counter_2020_up <- AC_inventory_counter %>% 
  filter(AC_inventory_counter$datetime_utc >= "2019-12-30 12:00:00 UTC", VoltageClassCodeName == "100-199 kV")

ggplot(AC_inventory_counter_2020_up, aes(x=datetime_utc, y = total_AC_Inv, color = VoltageClassCodeName)) +
  geom_line() +
  labs(
    title = "Hourly count of inventory",
    x = "Month", 
    y = "Inventory Count",
    color = "Voltage Class"  
  ) +
  theme_minimal() 

```

#Counting outages by duration and calculating outage rate
```{r}
TADS_AC_outages <- read_csv("/Users/JChuang/Documents/energymarkettool/outage_analysis/Processed/TADS_AC_Duration_Cleaned.csv")

#Filter outages list, add an outage end column, and convert times to UTC based on TimeZoneCode
outages_list <- TADS_AC_outages %>% 
  #filter(RegionCode == NERC_region) %>% 
  mutate(OutageEndDT = OutageStartDT + dhours(Outage_hrs)) %>% 
  mutate(
    TimeZoneOlson = tz_mapping[TimeZoneCode],  # create new column with Olson tz

    # Convert to local time using force_tz (treats input as naive time in local tz)
    OutageStart_local = force_tz(OutageStartDT, tzone = TimeZoneOlson),
    OutageEnd_local   = force_tz(OutageEndDT,   tzone = TimeZoneOlson),

    # Convert to UTC
    OutageStart_UTC = with_tz(OutageStart_local, tzone = "UTC"),
    OutageEnd_UTC   = with_tz(OutageEnd_local,   tzone = "UTC") )%>% 
  select(-OutageStart_local,-OutageEnd_local)

#Keeping only outages with duration >0 min
AC_outages_filtered <- outages_list %>% 
  filter(Outage_hrs > 0)

AC_outages_filter_copy <- AC_outages_filtered[!is.na(start) & !is.na(end)]
  
#Note: results from above code also used for next section

result_list <- list()

#convert to data.tables
setDT(AC_inventory_counter)
setDT(AC_outages_filtered)

#Creating interval columns for the join
AC_inventory_counter[, `:=`(hour_start = datetime_utc, hour_end = datetime_utc + lubridate::hours(1)- seconds(1))]

setnames(AC_outages_filtered, c("OutageStart_UTC", "OutageEnd_UTC"),
         c("start", "end"))  # rename for foverlaps

# Set keys for foverlaps
setkey(AC_outages_filtered, start, end)
setkey(AC_inventory_counter, hour_start, hour_end)

AC_outages_filtered <- AC_outages_filtered[!is.na(start) & !is.na(end)]

for (v in voltage_classes) {
  x_sub <- AC_inventory_counter[VoltageClassCodeName == v]
  y_sub <- AC_outages_filtered[VoltageClassCodeName == v]

  setkey(x_sub, hour_start, hour_end)
  setkey(y_sub, start, end)

  result_list[[v]] <- foverlaps(x_sub, y_sub, type = "any", nomatch = 0L) #drops rows in x_sub that have no matching overlap in y_sub. "any"
}

AC_outage_joined <- rbindlist(result_list)

#count outages per hour
AC_outage_counted <- AC_outage_joined %>% 
  group_by(datetime_utc, VoltageClassCodeName) %>% 
  summarise(num_outages_by_duration = n(),
            .groups = "drop")

#joining back with full counter_df
AC_outage_duration_counter <- AC_inventory_counter %>%
  left_join(AC_outage_counted, by = c("datetime_utc", "VoltageClassCodeName")) %>%
  mutate(num_outages_by_duration = replace_na(num_outages_by_duration, 0))

#cleaning up
AC_outage_duration_counter  <- AC_outage_duration_counter %>% 
  select(-c(hour_start, hour_end)) %>% 
  mutate(outage_rate = num_outages_by_duration/total_AC_Inv) 

ggplot(AC_outage_duration_counter, aes(x=datetime_utc, y = num_outages_by_duration, color = VoltageClassCodeName)) +
  geom_line() +
  labs(
    title = "Hourly count of outages",
    x = "Month", 
    y = "Outage Count",
    color = "Voltage Class"  
  ) +
  theme_minimal() 
```

#Counting outages by start

8/4 continue from here:
Make this work with what we have here! Then further below, join with other dataframe that has outage rate and inventory!
```{r}
# Get all unique voltage classes
voltage_classes <- AC_outages_filtered %>%
  distinct(VoltageClassCodeName) %>%
  pull()

# Create full hourly time grid
all_hours <- seq(
  from = floor_date(min(AC_outages_filter_copy$OutageStart_UTC, na.rm = TRUE), "hour"),
  to   = floor_date(max(AC_outages_filter_copy$OutageEnd_UTC, na.rm = TRUE), "hour"),
  by = "hour"
)

# Expand full grid
full_hourly_grid <- AC_inventory_counter %>%
  select(datetime_utc, VoltageClassCodeName) %>%
  distinct()

hourly_counts <- AC_outages_filter_copy %>%
  mutate(datetime_utc = floor_date(OutageStart_UTC, unit = "hour")) %>%
  count(VoltageClassCodeName, datetime_utc, name = "num_outages_by_start")

outage_start_counter <- full_hourly_grid %>%
  left_join(hourly_counts, by = c("VoltageClassCodeName", "datetime_utc")) %>%
  mutate(num_outages_by_start = replace_na(num_outages_by_start, 0))

ggplot(outage_start_counter, aes(x=datetime_utc, y = num_outages_by_start, color = VoltageClassCodeName)) +
  geom_line() +
  labs(
    title = "Hourly count of outage starts",
    x = "Month", 
    y = "Outage Start Count",
    color = "Voltage Class"  
  ) +
  theme_minimal() 
```

#Rate of outages caused by vegetation:

```{r}
# Create hourly counts by initiation cause
hourly_counts <- AC_outages_filter_copy %>%
  mutate(
    datetime_utc = floor_date(OutageStart_UTC, unit = "hour"),
    is_vegetation = grepl("vegetation", InitiationCauseCodeName, ignore.case = TRUE)
  ) %>%
  group_by(VoltageClassCodeName, datetime_utc) %>%
  summarise(
    total_outages = n(),
    vegetation_outages = sum(is_vegetation, na.rm = TRUE),
    vegetation_rate = vegetation_outages / total_outages,
    .groups = "drop"
  )

#plotting hourly counts of vegetation
ggplot(hourly_counts, aes(x = datetime_utc, y = vegetation_outages, color = VoltageClassCodeName)) +
  geom_line() +
  labs(
    title = "Hourly Vegetation Count",
    x = "Month",
    y = "Vegetation outage count",
    color = "Voltage Class"
  ) +
  facet_wrap(~ VoltageClassCodeName, scales = "free_y") +
  theme_minimal()

# Join with full hourly grid so missing hours show as 0
outage_start_counter <- full_hourly_grid %>%
  left_join(hourly_counts, by = c("VoltageClassCodeName", "datetime_utc")) %>%
  mutate(
    total_outages = replace_na(total_outages, 0),
    vegetation_outages = replace_na(vegetation_outages, 0),
    vegetation_rate = if_else(total_outages > 0,
                              vegetation_outages / total_outages,
                              NA_real_)   # NA where no outages occurred
  )

# Plot vegetation rate per hour
ggplot(outage_start_counter, aes(x = datetime_utc, y = vegetation_rate, color = VoltageClassCodeName)) +
  geom_line() +
  labs(
    title = "Hourly Vegetation Outage Rate",
    x = "Month",
    y = "Vegetation Outages / Total Outages",
    color = "Voltage Class"
  ) +
  facet_wrap(~ VoltageClassCodeName, scales = "free_y") +
  theme_minimal()
```
```{r}
# Calculate duration (in hours) for vegetation-caused outages
vegetation_durations <- AC_outages_filter_copy %>%
  filter(grepl("vegetation", InitiationCauseCodeName, ignore.case = TRUE)) %>%
  mutate(
    duration_hours = as.numeric(difftime(OutageEnd_UTC, OutageStart_UTC, units = "hours"))
  )

# Quick check of summary stats
vegetation_durations %>%
  summarise(
    avg_duration = mean(duration_hours, na.rm = TRUE),
    median_duration = median(duration_hours, na.rm = TRUE),
    max_duration = max(duration_hours, na.rm = TRUE)
  )

ggplot(vegetation_durations, aes(x = duration_hours)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  coord_cartesian(xlim = c(0, 72)) +   # zoom in to outages up to 3 days
  labs(
    title = "Histogram of Vegetation-Caused Outage Durations (0–72 hrs)",
    x = "Duration (hours)",
    y = "Number of Outages"
  ) +
  theme_minimal()

ggplot(vegetation_durations, aes(x = duration_hours)) +
  geom_histogram(binwidth = 0.2, fill = "steelblue", color = "white") +
  scale_x_log10() +
  labs(
    title = "Histogram of Vegetation-Caused Outage Durations (Log Scale)",
    x = "Duration (log hours)",
    y = "Number of Outages"
  ) +
  theme_minimal()

```

```{r}
# Row with maximum vegetation outage duration
max_row <- vegetation_durations %>%
  filter(!is.na(duration_hours)) %>%
  slice_max(duration_hours, n = 1)

max_row
```

