---
title: "01_saidi_saifi_caidi_duke"
output: html_document
---

```{r package install}
library(tidyverse)
library(lubridate)
library(here)
library(sf)
library(dplyr)
library(ggplot2)
library(readxl)
library(plotly)
library(shiny)
library(data.table)
library(tidycensus)
library(sf)
library(tigris)
library(jsonlite)
library(purrr)
library(plotly)
library(htmlwidgets)
library(shiny)
```

```{r Duke Energy outage data pull}
#outage_data_raw <- fread(here("data_raw/outage_tracker.csv"))
#saveRDS(outage_data_raw, here("data_processed/outage_tracker.rds"))
outage_data_raw <- readRDS(here("data_processed/outage_tracker.rds"))  #saving it in a binary format so it's easier to work with
```


Trying to extract county from device_lat and device_lon:
```{r adding spatial data}

# Cleans Duke Energy outage data, parses convex hulls,
# fills missing county/state, saves cleaned dataset
options(tigris_use_cache = TRUE)

# Make outages into sf object (points). Note: need to get rid of rows missing lat and lon though for this to work.
outages_sf <- outage_data_raw %>%
  filter(!is.na(device_lon), !is.na(device_lat)) %>%
  st_as_sf(coords = c("device_lon", "device_lat"), crs = 4326, remove = FALSE)

# Load county boundaries (for Duke’s states)
counties_sf <- counties(
  state = c("FL", "GA", "IL", "IN", "KY", "NC", "OH", "SC", "VA"),
  cb = TRUE
)

# Reproject outages to same CRS
outages_sf <- st_transform(outages_sf, crs = st_crs(counties_sf))

# Now the join works
outages_joined <- st_join(outages_sf, counties_sf, left = TRUE)

# Clean up
outages_final <- outages_joined %>%
  mutate(
    county = if_else(is.na(county), NAME, county),
    state  = if_else(is.na(state), STATEFP, state)
  )

```

SAIDI (System Average Interruption Duration Index)
= ∑(customers affected × outage duration) / total customers served

SAIFI (System Average Interruption Frequency Index)
= ∑(customers affected) / total customers served

CAIDI (Customer Average Interruption Duration Index)
= SAIDI / SAIFI

Ideal to use convex hulls, but unfortunately unsure of customers served in a convex hull, so going to stick with county data.

```{r tidycensus population data}
# population data by county for all the states

#Apply the API key
census_api_key("b5acdf7f791fa15dda8522426afc7a2e6e922c02", install = TRUE, overwrite = TRUE)
readRenviron("~/.Renviron")
```

```{r}
# From ACS (5-year estimates, includes households, income, etc.)
county_pop <- get_acs(
  geography = "county",
  variables = "B01003_001E",  # Total population
  year = 2022,
  state = c('FL', 'GA', 'IL', 'IN', 'KY', 'NC', 'OH', 'SC', 'VA')
)

county_pop_clean <- county_pop %>% 
  separate(NAME, into = c("NAMELSAD", "STATE_NAME"), sep = ", ") %>%
  rename(population = estimate) %>% 
  select("GEOID", "NAMELSAD", "STATE_NAME", "population")

outage_pop_combined <- left_join(outages_final, county_pop_clean, by = c('NAMELSAD', 'STATE_NAME'))

outage_pop_combined <- outage_pop_combined %>%
  # Drop old county/state columns if they exist
  select(-county, -state) %>%
  # Rename the shapefile fields
  rename(
    county = NAMELSAD,
    state  = STATE_NAME
  )
```

Creating new columns for the reliability metrics mentioned above:
```{r}
outage_pop_combined <- outage_pop_combined %>%
  mutate(
    # Extract numbers (replace missing with 0)
    days    = as.numeric(str_extract(fix_duration_estimate, "\\d+(?=d)")),
    hours   = as.numeric(str_extract(fix_duration_estimate, "\\d+(?=h)")),
    minutes = as.numeric(str_extract(fix_duration_estimate, "\\d+(?=m)")),
    seconds = as.numeric(str_extract(fix_duration_estimate, "\\d+(?=s)")),
    
    # Replace NA with 0
    across(c(days, hours, minutes, seconds), ~replace_na(., 0)),
    
    # Convert everything to minutes
    fix_duration_min = days * 24 * 60 + hours * 60 + minutes + seconds / 60
  )

reliability_metrics_table <- outage_pop_combined %>%
  group_by(state, county) %>%
  summarise(
    total_population = sum(population, na.rm = TRUE),
    total_affected   = sum(affected, na.rm = TRUE),
    total_duration   = sum(affected * fix_duration_min, na.rm = TRUE)
  ) %>%
  mutate(
    SAIDI = total_duration / total_population,
    SAIFI = total_affected / total_population,
    CAIDI = SAIDI / SAIFI
  ) %>%
  ungroup()
```
Pulling EIA-861 Data for state-level CAIDI, SAIFI, SAIDI for each state as of 2024
```{r}
# Read sheet (adjust sheet name if needed)
eia_reliability <- read_excel(
  "data_processed/Reliability_2024_Data_Early_Release_Clean.xlsx",
  sheet = "State Totals"
)

# Keep only state + metrics columns
eia_state_metrics <- eia_reliability %>%
  select(
    State,  # postal abbreviation like "NC", "FL"
    SAIDI_all   = `SAIDI (minutes per year) - All Events`,
    SAIFI_all   = `SAIFI (times per year) - All Events`,
    CAIDI_all   = `CAIDI (minutes per interruption) - All Events`,
    SAIDI_noMED = `SAIDI (minutes per year) - Without Major Event Days`,
    SAIFI_noMED = `SAIFI (times per year) - Without Major Event Days`,
    CAIDI_noMED = `CAIDI (minutes per interruption) - Without Major Event Days`
  ) %>%
  filter(!is.na(State)) %>%
  rename(state_code = State)  # rename for clarity

# Build state lookup
state_lookup <- fips_codes %>%
  select(state_code = state, state_name) %>%
  distinct()

# Attach full state name
eia_state_metrics <- eia_state_metrics %>%
  left_join(state_lookup, by = "state_code") %>% 
  rename(state = "state_name")

reliability_metrics_table_final <- left_join(reliability_metrics_table, eia_state_metrics, by = "state")

reliability_metrics_table_final <- reliability_metrics_table_final %>% 
  rename(SAIDI_county = SAIDI, SAIFI_county = SAIFI, CAIDI_county = CAIDI)

st_write(reliability_metrics_table_final, "./data_processed/reliability_metrics_table_final.gpkg")

```

Some visuals:

for each state, have three plots (CAIDI/SAIDI/SAIFI), bar for each county and then line showing location of the all and MED for state

```{r}
# Get unique states
all_states <- unique(reliability_metrics_table_final$state)

# Function to generate plots for one state
plot_reliability_state <- function(st) {
  df <- reliability_metrics_table_final %>% filter(state_code == st)
  
  p1 <- ggplot(df, aes(x = reorder(county, SAIDI_county), y = SAIDI_county)) +
    geom_col(fill = "steelblue") +
    geom_hline(aes(yintercept = SAIDI_all), color = "red", linetype = "dashed") +
    geom_hline(aes(yintercept = SAIDI_noMED), color = "darkgreen", linetype = "dotted") +
    coord_flip() +
    labs(title = paste("SAIDI by County (", st, ")", sep = ""), y = "SAIDI (minutes)", x = "County")
  
  p2 <- ggplot(df, aes(x = reorder(county, SAIFI_county), y = SAIFI_county)) +
    geom_col(fill = "steelblue") +
    geom_hline(aes(yintercept = SAIFI_all), color = "red", linetype = "dashed") +
    geom_hline(aes(yintercept = SAIFI_noMED), color = "darkgreen", linetype = "dotted") +
    coord_flip() +
    labs(title = paste("SAIFI by County (", st, ")", sep = ""), y = "SAIFI", x = "County")
  
  p3 <- ggplot(df, aes(x = reorder(county, CAIDI_county), y = CAIDI_county)) +
    geom_col(fill = "steelblue") +
    geom_hline(aes(yintercept = CAIDI_all), color = "red", linetype = "dashed") +
    geom_hline(aes(yintercept = CAIDI_noMED), color = "darkgreen", linetype = "dotted") +
    coord_flip() +
    labs(title = paste("CAIDI by County (", st, ")", sep = ""), y = "CAIDI (minutes)", x = "County")
  
  list(SAIDI = p1, SAIFI = p2, CAIDI = p3)
}

# Example: generate plots for Florida
plots_FL <- plot_reliability_state("FL")
plots_FL$SAIDI
plots_FL$SAIFI
plots_FL$CAIDI

plots_NC <- plot_reliability_state("NC")
plots_NC$SAIDI
plots_NC$SAIFI
plots_NC$CAIDI

```

```{r}
# Reuse the plotting logic from earlier but as a function that accepts df directly
plot_reliability_plotly_from_df <- function(df, metric = "SAIDI", n_pixel_per_row = 18, min_height = 400, left_margin = 220) {
  if (nrow(df) == 0) return(NULL)

  if (metric == "SAIDI") {
    val_col <- "SAIDI_county"; all_col <- "SAIDI_all"; noMED_col <- "SAIDI_noMED"; y_label <- "SAIDI (minutes)"
  } else if (metric == "SAIFI") {
    val_col <- "SAIFI_county"; all_col <- "SAIFI_all"; noMED_col <- "SAIFI_noMED"; y_label <- "SAIFI"
  } else {
    val_col <- "CAIDI_county"; all_col <- "CAIDI_all"; noMED_col <- "CAIDI_noMED"; y_label <- "CAIDI (minutes)"
  }

  df2 <- df %>%
    arrange(.data[[val_col]]) %>%
    mutate(county_ord = factor(county, levels = county))

  n <- nrow(df2)
  plot_height <- max(min_height, n * n_pixel_per_row + 120)

  df2 <- df2 %>%
    mutate(hover = paste0("<b>", county, "</b><br>",
                          y_label, ": ", round(.data[[val_col]], 3), "<br>",
                          "State (all): ", round(first(.data[[all_col]]), 3), "<br>",
                          "State (noMED): ", round(first(.data[[noMED_col]]), 3)))

  p <- plot_ly(data = df2,
               x = ~get(val_col),
               y = ~county_ord,
               type = "bar",
               orientation = "h",
               text = ~hover,
               hoverinfo = "text",
               marker = list(color = 'steelblue'),
               name = metric)

  state_all_val <- df2[[all_col]][1]
  state_noMED_val <- df2[[noMED_col]][1]

  # add benchmark vertical lines (x-lines)
  p <- p %>%
    add_lines(x = c(state_all_val, state_all_val),
              y = c(-0.5, n + 0.5),
              name = paste0(metric, " (state all)"),
              line = list(color = 'red', dash = 'dash'),
              hoverinfo = "text",
              text = paste0("State (all): ", round(state_all_val, 3)),
              inherit = FALSE, showlegend = TRUE) %>%
    add_lines(x = c(state_noMED_val, state_noMED_val),
              y = c(-0.5, n + 0.5),
              name = paste0(metric, " (state noMED)"),
              line = list(color = 'darkgreen', dash = 'dot'),
              hoverinfo = "text",
              text = paste0("State (noMED): ", round(state_noMED_val, 3)),
              inherit = FALSE, showlegend = TRUE)

  p <- p %>%
    layout(title = paste0(metric, " by County (", unique(df2$state_code), ")"),
           xaxis = list(title = y_label, zeroline = FALSE),
           yaxis = list(title = "", automargin = TRUE),
           margin = list(l = left_margin, r = 50, t = 60, b = 50),
           height = plot_height,
           hovermode = "closest")

  p
}

ui <- fluidPage(
  titlePanel("Reliability by County (interactive)"),
  sidebarLayout(
    sidebarPanel(
      selectInput("state", "State:", choices = sort(unique(reliability_metrics_table_final$state_code)), selected = head(sort(unique(reliability_metrics_table_final$state_code)), 1)),
      selectInput("metric", "Metric:", choices = c("SAIDI","SAIFI","CAIDI"), selected = "SAIDI"),
      sliderInput("pixel_per_row", "Pixels per row (vertical density):", min = 8, max = 40, value = 18),
      downloadButton("download_html", "Save current plot as HTML")
    ),
    mainPanel(
      plotlyOutput("plot", height = "auto")  # height will be set by plotly object
    )
  )
)

server <- function(input, output, session) {
  # reactive filtered df
  df_state <- reactive({
    req(input$state)
    reliability_metrics_table_final %>% filter(state_code == input$state)
  })

  output$plot <- renderPlotly({
    df <- df_state()
    req(nrow(df) > 0)
    plot_reliability_plotly_from_df(df, metric = input$metric, n_pixel_per_row = input$pixel_per_row)
  })

  output$download_html <- downloadHandler(
    filename = function() paste0("reliability_", input$state, "_", input$metric, ".html"),
    content = function(file) {
      p <- plot_reliability_plotly_from_df(df_state(), metric = input$metric, n_pixel_per_row = input$pixel_per_row)
      htmlwidgets::saveWidget(as_widget(p), file, selfcontained = TRUE)
    }
  )
}

shinyApp(ui, server)
```

